---
title: "2-Hour Microbiome Analysis Workshop: From Raw Reads to Biological Insight"
author: "Michael Sieler"
date: "Last updated: `r Sys.Date()`"
output:
  html_document:
    number_sections: true
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    code_folding: show
    df_print: paged
---

```{r setup, include=FALSE}
# Set default options for all code chunks in this document
knitr::opts_chunk$set(
  echo    = TRUE,     # Show code in the rendered output
  eval    = TRUE,     # Actually run the code (set FALSE to show but not execute)
  warning = FALSE,    # Suppress warning messages in the output
  message = FALSE,    # Suppress package loading messages in the output
  cache   = TRUE,     # Cache results so re-knitting is faster (skips unchanged chunks)
  dpi     = 150,      # Resolution for rendered figures (dots per inch)
  fig.width  = 8,     # Default figure width in inches
  fig.height = 5,     # Default figure height in inches
  out.width  = "95%"  # Scale figures to 95% of the page width in the HTML output
)

# Resolve project root to the directory containing the .Rproj file.
resolve_project_root <- function() {
  wd <- getwd()
  candidate_roots <- c(
    wd,
    file.path(wd, ".."),
    file.path(wd, "..", ".."),
    file.path(wd, "..", "..", "..")
  )
  candidate_roots <- unique(normalizePath(candidate_roots, winslash = "/", mustWork = FALSE))

  has_rproj <- vapply(
    candidate_roots,
    function(path) {
      file.exists(file.path(path, "MicrobiotaWorkshop_2026.Rproj"))
    },
    FUN.VALUE = logical(1)
  )

  if (!any(has_rproj)) {
    stop(
      "Could not locate tutorial project root from working directory: ", wd,
      ". Expected a parent directory containing 'MicrobiotaWorkshop_2026.Rproj'."
    )
  }

  candidate_roots[which(has_rproj)[1]]
}

project_root_dir <- resolve_project_root()

# Central output folders for tutorial artifacts.
results_base_dir <- file.path(project_root_dir, "Results")
dada2_output_dir <- file.path(results_base_dir, "DADA2")
microbiome_output_dir <- file.path(results_base_dir, "MicrobiomeAnalysis")
data_base_dir <- file.path(project_root_dir, "Data")

dada2_fig_dir <- file.path(dada2_output_dir, "Figures")
dada2_tbl_dir <- file.path(dada2_output_dir, "Tables")
microbiome_fig_dir <- file.path(microbiome_output_dir, "Figures")
microbiome_tbl_dir <- file.path(microbiome_output_dir, "Tables")

resolve_dataset_dir <- function(data_root, bioproject = "PRJNA792398") {
  candidates <- list.dirs(data_root, full.names = TRUE, recursive = FALSE)
  pattern <- paste0("^", bioproject, "__[0-9]{2}\\.[0-9]{2}\\.[0-9]{4}$")
  candidates <- candidates[grepl(pattern, basename(candidates))]

  if (length(candidates) == 0) {
    stop(
      "No dataset directory found under ", data_root,
      " matching ", bioproject, "__DD.MM.YYYY"
    )
  }

  candidates[which.max(file.info(candidates)$mtime)]
}

dir.create(dada2_fig_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(dada2_tbl_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(microbiome_fig_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(microbiome_tbl_dir, recursive = TRUE, showWarnings = FALSE)

# Part 1 figures default to Results/DADA2/Figures
knitr::opts_chunk$set(fig.path = file.path(dada2_fig_dir, "fig-"))
```

# PART 1: From Raw FastQ to ASV Table {.tabset}

## Introduction: What Are We Doing and Why?

Microbiome studies aim to characterize the microbial communities living in or on
a host, environment, or ecosystem. The **16S ribosomal RNA gene** is the most
widely used phylogenetic marker for bacteria and archaea because it contains both
highly conserved regions (enabling universal primer binding) and hypervariable
regions (V1--V9) that provide taxonomic discrimination.

In this workshop, we will process raw Illumina paired-end sequencing data through
the **DADA2** pipeline to generate a table of **Amplicon Sequence Variants
(ASVs)** with taxonomy assignments, then analyze the resulting communities.

**The pipeline at a glance:**

```
Raw paired-end reads (FastQ)
    │
    ▼
Quality inspection ──► Primer removal (cutadapt / DADA2 trimLeft)
    │
    ▼
Quality filtering (DADA2 filterAndTrim)
    │
    ▼
Error model learning ──► Denoising (ASV inference)
    │
    ▼
Paired-read merging ──► Chimera removal
    │
    ▼
ASV table ──► Taxonomy assignment (SILVA)
    │
    ▼
Contaminant identification (decontam)
    │
    ▼
Taxa filtering (non-target, unresolved, low-prevalence)
    │
    ▼
Phyloseq object
```

### ASVs vs. OTUs

Historically, amplicon sequences were clustered into **Operational Taxonomic
Units (OTUs)** at 97% identity---a pragmatic but arbitrary threshold that
masks real biological variation and is not reproducible across studies.

**ASVs** use statistical error models to resolve exact biological sequences.
They are:

- **Higher resolution** --- can distinguish sequences differing by a single
  nucleotide
- **Reproducible** --- the same sequence gets the same ASV identity in any study
- **Comparable across studies** --- no need for re-clustering

DADA2 (Callahan et al., 2016, *Nature Methods*) is the most widely used ASV
inference tool and is what we use in this workshop.

### Dataset

We are working with **honeybee 16S V4 amplicon data** (515F--806R primers,
Illumina MiSeq). [TODO: Describe the dataset — experimental design, sample
groups, replicates, and any relevant biological context.]

### Setup

```{r load-libraries-part1}
# dada2: the main pipeline tool that takes raw sequencing reads and infers
#   exact biological sequences (ASVs) by modeling sequencing errors
library(dada2)

# phyloseq: bundles your ASV table, taxonomy, and sample metadata into a
#   single object for easy manipulation and analysis
library(phyloseq)

# decontam: identifies likely contaminant sequences introduced from lab
#   reagents, kits, or the environment (not from your biological samples)
library(decontam)

# RColorBrewer: colorblind-friendly palettes for discrete/continuous plotting
library(RColorBrewer)

# tidyverse: a collection of R packages (dplyr, ggplot2, tidyr, etc.) for
#   data wrangling, transformation, and visualization
library(tidyverse)

# Number of CPU cores DADA2 will use for parallelized steps (e.g., error
# learning, denoising). Increase this if your machine has more cores to
# speed up computation. Decrease if you run into memory issues.
# Windows VMs are more stable with single-core processing for this workflow.
# On macOS/Linux, use up to 10 cores (bounded by available cores).
threads <- if (.Platform$OS.type == "windows") {
  1L
} else {
  max(1L, min(10L, as.integer(parallel::detectCores(logical = FALSE))))
}

# Normalize sample IDs once, early, to avoid accidental "X" prefixes that can
# appear when IDs begin with numbers (e.g., X01 -> 01).
normalize_sample_id <- function(x) {
  x <- as.character(x)
  x <- sub("^X(?=[0-9])", "", x, perl = TRUE)
  ifelse(grepl("^[0-9]+$", x), paste0("Bee_", x), x)
}

# Checkpoint folders for intermediate RDS files under Data/.
dada2_rds_dir <- file.path(data_base_dir, "DADA2")
microbiome_rds_dir <- file.path(data_base_dir, "MicrobiomeAnalysis")
dada2_intermediate_dir <- file.path(dada2_rds_dir, "intermediate_fastq")
dir.create(dada2_rds_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(microbiome_rds_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(dada2_intermediate_dir, recursive = TRUE, showWarnings = FALSE)

# Workshop default: load precomputed DADA2 checkpoints instead of recomputing.
use_precomputed_dada2 <- TRUE

require_rds_checkpoint <- function(path, label) {
  if (!file.exists(path)) {
    stop(
      "Required precomputed checkpoint missing for ", label, ": ", path,
      "\nEither provide this .rds file or set use_precomputed_dada2 <- FALSE to recompute."
    )
  }
  readRDS(path)
}
```

> **Note on computation time:** Several DADA2 steps (error learning, denoising,
> taxonomy assignment) take 10--40 minutes on real datasets. For this workshop,
> we provide **pre-computed results** as `.rds` files. The actual function calls
> are shown but commented out. You can uncomment and run them on your own time.

------------------------------------------------------------------------

## Inspect Raw Data and Quality

The first step in any sequencing analysis is to examine the raw data. We need to
understand: how many reads do we have, what is their quality, and are there any
non-biological sequences (primers, adapters) that need to be removed?

### List the input files

Our raw data consists of **paired-end reads**: forward (R1) and reverse (R2)
files for each sample, in gzipped FastQ format.

```{r list-input-files}
# Paths to the tutorial subset data generated in the data setup step.
dataset_dir <- resolve_dataset_dir(data_base_dir, "PRJNA792398")
metadata_path <- file.path(data_base_dir, "Metadata", "GutMicrobiota_BeeTracking_metadata__TutorialSubset.csv")
samplesheet_path <- file.path(dataset_dir, "samplesheet__TutorialSubset.csv")
raw_dir <- file.path(dataset_dir, "FASTQs", "fastq_raw")
symlink_dir <- file.path(dataset_dir, "FASTQs", "fastq_symlinks__TutorialSubset")

# Read metadata and ensure expected ID column exists.
metadata_tbl <- read.csv(metadata_path, stringsAsFactors = FALSE, check.names = FALSE)
stopifnot("Sample_ID" %in% colnames(metadata_tbl))
metadata_tbl$Sample_ID <- normalize_sample_id(metadata_tbl$Sample_ID)

# Build a paired file table from the tutorial samplesheet.
# Prefer raw FASTQ files (Windows VM friendly), and fallback to symlink paths when needed.
if (!file.exists(samplesheet_path)) {
  stop("Tutorial samplesheet not found: ", samplesheet_path)
}

samplesheet_map <- read.csv(samplesheet_path, stringsAsFactors = FALSE, check.names = FALSE)
stopifnot(all(c("sample_id", "run_accession", "forward", "reverse") %in% colnames(samplesheet_map)))

samplesheet <- samplesheet_map %>%
  dplyr::mutate(
    sample_id = normalize_sample_id(.data$sample_id),
    forward_raw = file.path(raw_dir, paste0(.data$run_accession, "_1.fastq.gz")),
    reverse_raw = file.path(raw_dir, paste0(.data$run_accession, "_2.fastq.gz")),
    forward_symlink = ifelse(
      grepl("^([A-Za-z]:|/)", .data$forward),
      .data$forward,
      file.path(project_root_dir, .data$forward)
    ),
    reverse_symlink = ifelse(
      grepl("^([A-Za-z]:|/)", .data$reverse),
      .data$reverse,
      file.path(project_root_dir, .data$reverse)
    )
  ) %>%
  dplyr::mutate(
    use_raw = file.exists(.data$forward_raw) & file.exists(.data$reverse_raw),
    forward = ifelse(.data$use_raw, .data$forward_raw, .data$forward_symlink),
    reverse = ifelse(.data$use_raw, .data$reverse_raw, .data$reverse_symlink)
  ) %>%
  dplyr::filter(.data$sample_id %in% metadata_tbl$Sample_ID) %>%
  dplyr::arrange(.data$sample_id) %>%
  dplyr::select("sample_id", "run_accession", "forward", "reverse")

if (anyDuplicated(samplesheet$sample_id) > 0) {
  dup_ids <- unique(samplesheet$sample_id[duplicated(samplesheet$sample_id)])
  stop(
    "Duplicate sample IDs detected after building the tutorial samplesheet. ",
    "Examples: ", paste(utils::head(dup_ids, 8), collapse = ", ")
  )
}

if (!all(file.exists(samplesheet$forward)) || !all(file.exists(samplesheet$reverse))) {
  missing_f <- samplesheet$forward[!file.exists(samplesheet$forward)]
  missing_r <- samplesheet$reverse[!file.exists(samplesheet$reverse)]
  stop(
    "Missing FASTQ file(s) after path resolution. ",
    "Examples: ",
    paste(utils::head(c(missing_f, missing_r), 6), collapse = ", ")
  )
}

# Build ordered file vectors for DADA2.
fnFs <- samplesheet$forward
fnRs <- samplesheet$reverse
sample.names <- samplesheet$sample_id
names(fnFs) <- sample.names
names(fnRs) <- sample.names

# Quick sanity check: how many samples did we find?
cat("Number of samples:", length(sample.names), "\n")
cat("First few sample names:", head(sample.names), "\n")
cat("FASTQ raw directory:", raw_dir, "\n")
cat("FASTQ symlink directory:", symlink_dir, "\n")
cat("Tutorial samplesheet:", samplesheet_path, "\n")
cat("Metadata file:", metadata_path, "\n")
```

### Quality profiles

The `plotQualityProfile()` function shows the distribution of quality scores at
each position across all reads. This is critical for deciding where to trim.

**What to look for:**

- The green line shows mean quality per position
- The orange lines show quartiles of the quality score distribution
- Quality typically degrades toward the 3' end (right side) because Illumina's
  sequencing-by-synthesis chemistry accumulates phasing errors over cycles
- R2 (reverse) reads are generally lower quality than R1 (forward)

```{r quality-profiles, fig.height=4, fig.width=10}
# plotQualityProfile() reads the quality scores from all FastQ files and
# creates a summary plot showing how quality changes along the length of reads.
# aggregate = TRUE combines all samples into one plot (instead of one per sample).
# This helps us decide WHERE to trim reads: we want to cut off the low-quality tails.
# Force serial QA workers on Windows to avoid socket/serialize warnings.
quality_bpparam <- if (.Platform$OS.type == "windows") {
  BiocParallel::SerialParam()
} else {
  BiocParallel::MulticoreParam(workers = threads)
}
old_bpparam <- BiocParallel::bpparam()
BiocParallel::register(quality_bpparam)

quality_F <- dada2::plotQualityProfile(fnFs, aggregate = TRUE) +
  ggplot2::ggtitle("Forward reads (R1)")  # Add a descriptive title
quality_R <- dada2::plotQualityProfile(fnRs, aggregate = TRUE) +
  ggplot2::ggtitle("Reverse reads (R2)")
BiocParallel::register(old_bpparam)

# Display the plots — look for where quality drops below ~Q20-Q30
quality_F
quality_R
```

> **Quality score refresher:**
>
> - Q30 = 1 error per 1,000 bases (99.9% accuracy)
> - Q20 = 1 error per 100 bases (99% accuracy)
> - Q10 = 1 error per 10 bases (90% accuracy)
>
> As a rule of thumb, trimming at positions where quality drops below Q20--Q30
> is standard practice.

------------------------------------------------------------------------

## Primer Detection and Quality Filtering

### Checking for primer sequences

Depending on the library preparation protocol, primer sequences may or may not
be present at the 5' end of reads. **If they are present, they must be
removed** because:

- Primers are synthetic sequences, not biological signal
- Degenerate bases in primers (e.g., `[AC]` meaning A or C) create artificial
  sequence variants
- Primer-target mismatches would inflate apparent diversity

We are using the **515F--806R (Caporaso) primers** targeting the V4 region:

```{r define-primers}
# Primers are short synthetic DNA sequences used during PCR to target a specific
# gene region (here, the V4 region of the 16S rRNA gene).
# Brackets like [AC] indicate "degenerate" bases — the primer can have either
# A or C at that position, allowing it to bind a broader range of bacteria.
# We define both the forward primer and its reverse complement (.rc), because
# primers can appear in either orientation depending on read direction.

# 515F forward primer (19 bp) — binds the start of the V4 region
R1_primer    <- "GTGCCAGC[AC]GCCGCGGTAA"
# Reverse complement of the forward primer
R1_primer.rc <- "TTACCGCGGC[TG]GCTGGCAC"

# 806R reverse primer (20 bp) — binds the end of the V4 region
R2_primer    <- "GGACTAC[ACT][CGT]GGGT[AT]TCTAAT"
# Reverse complement of the reverse primer
R2_primer.rc <- "ATTAGA[AT]ACCC[CGT][AGT]GTAGTCC"
```

```{r check-primers}
# Before removing primers, we first check whether they are actually present
# in our reads. We test one sample as a representative.

# getSequences() reads all DNA sequences from a FastQ file into a character vector
test_seqs_R1 <- dada2::getSequences(fnFs[1])
cat("Total reads in test file:", length(test_seqs_R1), "\n\n")

# str_locate() searches each read for the primer pattern (with regex support
# for degenerate bases). It returns the start/end position of the match, or
# NA if no match is found. We count how many reads contain the primer.
R1_matches <- data.frame(stringr::str_locate(test_seqs_R1, R1_primer))
cat("R1 reads with forward primer detected:",
    sum(!is.na(R1_matches$start)), "of", length(test_seqs_R1),
    paste0("(", round(sum(!is.na(R1_matches$start)) / length(test_seqs_R1) * 100, 1), "%)"),
    "\n")

# Repeat the same primer check for the reverse reads (R2)
test_seqs_R2 <- dada2::getSequences(fnRs[1])
R2_matches <- data.frame(stringr::str_locate(test_seqs_R2, R2_primer))
cat("R2 reads with reverse primer detected:",
    sum(!is.na(R2_matches$start)), "of", length(test_seqs_R2),
    paste0("(", round(sum(!is.na(R2_matches$start)) / length(test_seqs_R2) * 100, 1), "%)"),
    "\n")
```

> The primers are present in most reads, so we need to remove them before
> proceeding with the pipeline.

### Primer removal: Two approaches

There are two common strategies for removing primers from amplicon reads. The
choice depends on your library preparation and how consistently primers appear:

**Approach 1: Cutadapt (external tool, recommended for most cases)**

[Cutadapt](https://cutadapt.readthedocs.io/) is a dedicated primer/adapter
removal tool run on the command line *before* DADA2. It is the more robust
option because it:

- Handles **degenerate bases** in primers natively
- Can find and remove primers even when they don't start at exactly position 1
  (e.g., due to indels or sequencing offsets)
- Trims adapters from both the 5' and 3' ends, including read-through into
  the opposite primer
- Reports per-sample trimming statistics

```{bash cutadapt-example, eval=FALSE}
# Example cutadapt command for paired-end primer removal.
# Cutadapt is a command-line tool (not R) that finds and removes primer/adapter
# sequences from reads. This loop processes each sample's paired-end files.
#
# Flag reference:
#   -g : forward primer sequence to remove from the 5' end of R1 reads
#   -G : reverse primer sequence to remove from the 5' end of R2 reads
#   -o : output file for trimmed R1 reads
#   -p : output file for trimmed R2 reads (the "paired" output)
#   --pair-filter=any : if EITHER read in a pair fails, discard the whole pair

# Loop over every R1 file in the FASTQ symlink folder
mkdir -p cutadapt
for f in Data/PRJNA792398__*/FASTQs/fastq_symlinks/*_1.fastq.gz; do
  r=${f/_1.fastq.gz/_2.fastq.gz}   # Derive matching reverse-read filename
  # Build output filenames in a "cutadapt/" folder, adding ".cut" to distinguish them
  out1=cutadapt/$(basename $f .fastq.gz).cut.fastq.gz
  out2=cutadapt/$(basename $r .fastq.gz).cut.fastq.gz
  # Run cutadapt on paired-end reads.
  # IMPORTANT: in bash, a line ending in "\" cannot have trailing spaces/comments.
  cutadapt \
    -g GTGCCAGCMGCCGCGGTAA \
    -G GGACTACHVGGGTWTCTAAT \
    -o "$out1" -p "$out2" \
    --pair-filter=any \
    "$f" "$r"
done
```

> **When to use cutadapt:** If your library prep uses the two-step Illumina
> protocol (primers in the read), cutadapt is the gold standard. If you are
> using the EMP protocol (primers removed by sequencing primers), cutadapt is
> usually unnecessary.

**Approach 2: DADA2 `trimLeft` (simpler, used in this tutorial)**

When primers always start at position 1 (as is the case for our dataset),
DADA2's `filterAndTrim()` can remove them via the `trimLeft` parameter. This
is simpler (no external tool), but less flexible---it removes a fixed number of
bases from the 5' end regardless of whether a primer is present.

For this tutorial, we use `trimLeft` since our primers consistently start at
position 1 (as confirmed above).

### Quality filtering and trimming

`filterAndTrim()` is the workhorse function that handles multiple tasks at once.
Each parameter serves a specific purpose:

| Parameter | Value | Purpose |
|-----------|-------|---------|
| `trimLeft` | `c(19, 20)` | Remove 19 bp (F primer) and 20 bp (R primer) from 5' end |
| `truncLen` | `0` (no trim) | Trim low-quality 3' bases; set based on quality profiles. `0` = no 3' trimming (our data is high quality throughout) |
| `maxN` | `0` | Discard reads with any ambiguous bases (N). DADA2 requires this |
| `maxEE` | `c(2, 2)` | Maximum expected errors per read. A read with maxEE=2 has on average ≤2 incorrect bases |
| `truncQ` | `2` | Truncate reads at the first base with quality score ≤ 2 (additional safeguard against very low-quality tails) |
| `rm.phix` | `TRUE` | Remove PhiX spike-in sequences (added for sequencing complexity on amplicon runs) |

```{r filter-and-trim}
# Remove stale filtered outputs from older naming schemes (for example 01_*, Bee01_*),
# so this run keeps only the current sample-ID convention in intermediate_fastq.
if (!use_precomputed_dada2) {
  old_filtered_files <- list.files(
    dada2_intermediate_dir,
    pattern = "_[FR]_filt\\.fastq\\.gz$",
    full.names = TRUE
  )
  if (length(old_filtered_files) > 0) {
    file.remove(old_filtered_files)
  }
}

# Create output file paths for the filtered reads — one forward and one
# reverse file per sample, stored in a "filtered" subfolder
filtFs <- file.path(dada2_intermediate_dir, paste0(sample.names, "_F_filt.fastq.gz"))
filtRs <- file.path(dada2_intermediate_dir, paste0(sample.names, "_R_filt.fastq.gz"))

# Name each path with its sample ID so we can look up files by sample name later
names(filtFs) <- sample.names
names(filtRs) <- sample.names
cat("Filtered FASTQ output directory:", dada2_intermediate_dir, "\n")

# filterAndTrim() does several things at once:
#   - Removes primer sequences from the 5' (left) end of each read
#   - Discards reads that are too low-quality or contain ambiguous bases
#   - Optionally truncates the 3' (right) end at a fixed length
#   - Removes PhiX spike-in sequences (a control DNA added during sequencing)
# It writes new, cleaned FastQ files to Data/DADA2/intermediate_fastq.
rds_out <- file.path(dada2_rds_dir, "01_filterAndTrim_out.rds")
if (use_precomputed_dada2) {
  out <- require_rds_checkpoint(rds_out, "filterAndTrim output")
  message("Loaded precomputed filterAndTrim checkpoint: ", rds_out)
} else if (file.exists(rds_out)) {
  out <- readRDS(rds_out)
  message("Loaded precomputed filterAndTrim checkpoint: ", rds_out)
} else {
  out <- dada2::filterAndTrim(
    fwd      = fnFs,       # Input forward read files
    filt     = filtFs,     # Output (filtered) forward read files
    rev      = fnRs,       # Input reverse read files
    filt.rev = filtRs,     # Output (filtered) reverse read files
    trimLeft  = c(19, 20), # Remove 19 bp from forward reads (primer) and 20 bp from reverse
    truncLen  = 0,         # 0 = don't truncate the 3' end (our quality stays high throughout)
    maxN      = 0,         # Discard any read with ambiguous bases ("N"s); DADA2 requires this
    maxEE     = c(2, 2),   # Max "expected errors" per read; acts as a quality filter
    truncQ    = 2,         # Truncate read at the first base with quality score <= 2
    rm.phix   = TRUE,      # Remove PhiX control sequences (a common sequencing spike-in)
    compress  = TRUE,      # Write output files as compressed .gz to save disk space
    multithread = threads  # Use multiple CPU cores to speed up processing
  )
  saveRDS(out, rds_out)
}
utils::write.csv(
  as.data.frame(out),
  file.path(dada2_tbl_dir, "01_filterAndTrim_summary.csv"),
  row.names = TRUE
)
# Optional fast restart (uncomment if this step is too slow):
# if (file.exists(file.path(dada2_rds_dir, "01_filterAndTrim_out.rds"))) {
#   out <- readRDS(file.path(dada2_rds_dir, "01_filterAndTrim_out.rds"))
# }

# Check if any samples lost ALL their reads during filtering (column 2 = reads out)
# This can happen if a sample had extremely poor quality
zero_reads <- sample.names[out[, 2] == 0]
if (length(zero_reads) > 0) {
  cat("WARNING:", length(zero_reads), "sample(s) lost ALL reads during filtering:\n")
  cat("  ", paste(zero_reads, collapse = ", "), "\n")
}

# Update our file lists to only include samples that still have reads
# (file.exists checks that a filtered file was actually created)
if (!use_precomputed_dada2) {
  filtFs <- filtFs[file.exists(filtFs)]
  filtRs <- filtRs[file.exists(filtRs)]
}
cat("Samples with reads after filtering:", length(filtFs), "\n\n")

# Display the filtering summary table: columns are "reads.in" and "reads.out"
out
```

### Verify primer removal

After trimming, we confirm that primers have been successfully removed:

```{r verify-primer-removal}
# After filtering, we verify that trimLeft successfully removed the primer
# sequences. We repeat the same primer search on the FILTERED reads.
# If trimming worked, we expect zero matches.
if (length(filtFs) == 0 || length(filtRs) == 0 || !file.exists(filtFs[1]) || !file.exists(filtRs[1])) {
  cat("Skipping primer-removal verification: filtered FASTQ files are not available in this run mode.\n")
} else {
  # Read the first filtered forward file and search for the forward primer
  test_filt_R1 <- dada2::getSequences(filtFs[1])
  R1_filt_matches <- data.frame(stringr::str_locate(test_filt_R1, R1_primer))
  cat("Primers remaining in filtered R1 reads:", sum(!is.na(R1_filt_matches$start)), "\n")

  # Same check for the filtered reverse reads
  test_filt_R2 <- dada2::getSequences(filtRs[1])
  R2_filt_matches <- data.frame(stringr::str_locate(test_filt_R2, R2_primer))
  cat("Primers remaining in filtered R2 reads:", sum(!is.na(R2_filt_matches$start)), "\n")
}
```

> Zero primer matches in the filtered reads confirms successful removal.

------------------------------------------------------------------------

## Error Rate Learning and Denoising

This is the core of the DADA2 algorithm and what makes ASV inference possible.

### Learning error rates

DADA2 models the probability of every possible nucleotide transition (A→C, A→G,
A→T, etc.) as a function of the quality score at that position. This model is
learned **empirically from the data itself** by examining ~100 million+ bases.

```{r learn-errors}
# learnErrors() is the foundation of DADA2's denoising approach. It examines
# millions of base calls and their quality scores to learn the probability
# that each type of sequencing error occurs (e.g., an A being misread as a C).
# This error model is what allows DADA2 to distinguish REAL biological
# differences between sequences from ERRORS introduced by the sequencer.
#
# learnErrors() takes ~30 minutes on real data.
rds_errF <- file.path(dada2_rds_dir, "02_errF.rds")
rds_errR <- file.path(dada2_rds_dir, "03_errR.rds")
if (use_precomputed_dada2) {
  errF <- require_rds_checkpoint(rds_errF, "forward error model")
  errR <- require_rds_checkpoint(rds_errR, "reverse error model")
  message("Loaded precomputed error models: ", basename(rds_errF), ", ", basename(rds_errR))
} else if (file.exists(rds_errF) && file.exists(rds_errR)) {
  errF <- readRDS(rds_errF)
  errR <- readRDS(rds_errR)
  message("Loaded precomputed error models: ", basename(rds_errF), ", ", basename(rds_errR))
} else {
  set.seed(42)   # Ensures reproducibility when randomize = TRUE
  errF <- dada2::learnErrors(
    filtFs,             # Filtered forward read files
    nbases = 1e8,       # Use 100 million bases for learning (enough for a robust model)
    multithread = threads,  # Parallelize across CPU cores
    randomize = TRUE    # Randomly pick samples instead of using the first N in order
  )
  set.seed(42)
  errR <- dada2::learnErrors(
    filtRs,             # Filtered reverse read files
    nbases = 1e8,
    multithread = threads,
    randomize = TRUE
  )
  saveRDS(errF, rds_errF)
  saveRDS(errR, rds_errR)
}
# Optional fast restart (uncomment if this step is too slow):
# if (file.exists(file.path(dada2_rds_dir, "02_errF.rds"))) {
#   errF <- readRDS(file.path(dada2_rds_dir, "02_errF.rds"))
# }
# if (file.exists(file.path(dada2_rds_dir, "03_errR.rds"))) {
#   errR <- readRDS(file.path(dada2_rds_dir, "03_errR.rds"))
# }
```

> **Key parameters:** `nbases = 1e8` limits learning to ~100 million bases
> (sufficient for a robust model). `randomize = TRUE` randomly selects samples
> for learning instead of using the first N (avoids bias if samples are ordered
> by condition). `set.seed(42)` ensures reproducibility of the random selection.

### Visualize the error model

The error rate plots are essential quality control. Each panel shows one
nucleotide transition (e.g., A→C):

- **Points:** Observed error rates at each quality score
- **Black line:** Estimated error rates (the learned model)
- **Red line:** Expected error rates under nominal Q-scores

The black line should approximately follow the points and show a clear
downward trend (higher quality = fewer errors). This is the key innovation
enabling ASV resolution: instead of arbitrarily collapsing similar sequences
at 97%, DADA2 uses a principled statistical model to distinguish real biological
variation from sequencing errors.

```{r plot-errors, fig.height=6, fig.width=10}
# plotErrors() visualizes the learned error model. Each panel shows one type
# of nucleotide substitution (e.g., A→C, A→G). The x-axis is quality score
# and the y-axis is the error rate. The black line (learned model) should
# follow the observed data points and decrease as quality increases.
# nominalQ = TRUE overlays the red line showing the "expected" error rate
# based on quality scores alone (before DADA2's correction).
plotFerr <- dada2::plotErrors(errF, nominalQ = TRUE)

# suppressWarnings() hides minor ggplot warnings that don't affect the plot
suppressWarnings(print(plotFerr))
```

### Dereplication, denoising, and merging

These three steps are performed together in a per-sample loop to manage memory:

1. **Dereplication** (`derepFastq()`): Collapses identical reads, keeping track
   of abundance. This speeds up computation without losing information.

2. **Denoising** (`dada()`): The core ASV inference step. For each unique
   sequence, DADA2 evaluates whether it is likely a real biological variant or a
   sequencing error of a more abundant sequence, using the learned error model.

3. **Merging** (`mergePairs()`): Joins denoised forward and reverse reads based
   on their overlapping region. Default minimum overlap = 12 bp. Reads with
   mismatches in the overlap are discarded.

```{r denoise-and-merge}
# Pre-allocate empty lists/vectors to store results for each sample
rds_dadaFs <- file.path(dada2_rds_dir, "04_dadaFs.rds")
rds_dadaRs <- file.path(dada2_rds_dir, "05_dadaRs.rds")
rds_mergers <- file.path(dada2_rds_dir, "06_mergers.rds")
if (use_precomputed_dada2) {
  dadaFs <- require_rds_checkpoint(rds_dadaFs, "forward denoised reads")
  dadaRs <- require_rds_checkpoint(rds_dadaRs, "reverse denoised reads")
  mergers <- require_rds_checkpoint(rds_mergers, "merged pairs")
  sample.names.filt <- sort(intersect(names(dadaFs), names(mergers)))
  if (length(sample.names.filt) == 0) {
    stop("No overlapping sample names found between precomputed dadaFs and mergers checkpoints.")
  }
  message("Loaded precomputed denoise/merge checkpoints.")
} else if (file.exists(rds_dadaFs) && file.exists(rds_dadaRs) && file.exists(rds_mergers)) {
  dadaFs <- readRDS(rds_dadaFs)
  dadaRs <- readRDS(rds_dadaRs)
  mergers <- readRDS(rds_mergers)
  sample.names.filt <- sort(intersect(names(dadaFs), names(mergers)))
  if (length(sample.names.filt) == 0) {
    stop("No overlapping sample names found between loaded dadaFs and mergers checkpoints.")
  }
  message("Loaded precomputed denoise/merge checkpoints.")
} else {
  # This loop processes each sample through dereplication, denoising, and merging.
  sample.names.filt <- names(filtFs)
  mergers     <- vector("list", length(sample.names.filt))   # Merged read pairs
  ddFs_counts <- numeric(length(sample.names.filt))           # Forward read counts after denoising
  ddRs_counts <- numeric(length(sample.names.filt))           # Reverse read counts after denoising
  names(mergers)     <- sample.names.filt
  names(ddFs_counts) <- sample.names.filt
  names(ddRs_counts) <- sample.names.filt
  dadaFs <- vector("list", length(sample.names.filt))  # Forward denoising results
  dadaRs <- vector("list", length(sample.names.filt))  # Reverse denoising results
  names(dadaFs) <- sample.names.filt
  names(dadaRs) <- sample.names.filt

  # Process one sample at a time to keep memory usage manageable
  for (sam in sample.names.filt) {
    message("Processing sample: ", sam)
    derepF <- dada2::derepFastq(filtFs[[sam]])
    ddF    <- dada2::dada(derepF, err = errF, multithread = threads)
    derepR <- dada2::derepFastq(filtRs[[sam]])
    ddR    <- dada2::dada(derepR, err = errR, multithread = threads)
    merger <- dada2::mergePairs(ddF, derepF, ddR, derepR, verbose = TRUE)

    mergers[[sam]] <- merger
    dadaFs[[sam]]  <- ddF
    dadaRs[[sam]]  <- ddR
    ddFs_counts[sam] <- sum(dada2::getUniques(ddF))
    ddRs_counts[sam] <- sum(dada2::getUniques(ddR))
    rm(derepF, derepR, ddF, ddR, merger)
  }

  saveRDS(dadaFs, rds_dadaFs)
  saveRDS(dadaRs, rds_dadaRs)
  saveRDS(mergers, rds_mergers)
}
# Optional fast restart (uncomment if this step is too slow):
# if (file.exists(file.path(dada2_rds_dir, "04_dadaFs.rds"))) {
#   dadaFs <- readRDS(file.path(dada2_rds_dir, "04_dadaFs.rds"))
# }
# if (file.exists(file.path(dada2_rds_dir, "05_dadaRs.rds"))) {
#   dadaRs <- readRDS(file.path(dada2_rds_dir, "05_dadaRs.rds"))
# }
# if (file.exists(file.path(dada2_rds_dir, "06_mergers.rds"))) {
#   mergers <- readRDS(file.path(dada2_rds_dir, "06_mergers.rds"))
# }
```

------------------------------------------------------------------------

## ASV Table, Chimera Removal, and Taxonomy

### Build the sequence table

The sequence table is the samples-by-ASV count matrix. Column names are the
actual DNA sequences of each ASV.

```{r sequence-table}
# makeSequenceTable() combines the merged reads from ALL samples into a single
# count matrix: rows = samples, columns = unique ASV sequences.
# Each cell contains the number of times that ASV was observed in that sample.
# The column names are the actual DNA sequences of each ASV.
rds_seqtab <- file.path(dada2_rds_dir, "07_seqtab.rds")
if (use_precomputed_dada2) {
  seqtab <- require_rds_checkpoint(rds_seqtab, "sequence table")
  message("Loaded precomputed sequence table checkpoint: ", rds_seqtab)
} else if (file.exists(rds_seqtab)) {
  seqtab <- readRDS(rds_seqtab)
  message("Loaded precomputed sequence table checkpoint: ", rds_seqtab)
} else {
  seqtab <- dada2::makeSequenceTable(mergers)
  saveRDS(seqtab, rds_seqtab)
}
# Optional fast restart (uncomment if this step is too slow):
# if (file.exists(file.path(dada2_rds_dir, "07_seqtab.rds"))) {
#   seqtab <- readRDS(file.path(dada2_rds_dir, "07_seqtab.rds"))
# }

# dim() returns rows (samples) and columns (ASVs)
cat("Dimensions (samples x ASVs):", dim(seqtab), "\n\n")

# Check the length distribution of our merged sequences.
# nchar() counts characters (bases) in each ASV sequence.
# For V4 amplicons, we expect ~251-255 bp. Sequences far outside this
# range are likely artifacts (e.g., non-specific amplification or failed merges).
cat("Sequence length distribution:\n")
table(nchar(dada2::getSequences(seqtab)))
```

### Length filtering

For the V4 region (515F--806R), we expect merged amplicons of ~251--255 bp
(806 - 515 = 291, minus primers: 291 - 19 - 20 ≈ 252, with some biological
variation from indels). Sequences far outside this range likely represent
non-specific amplification or merging artifacts.

```{r length-filter}
# Keep only ASV sequences whose length falls within the expected range (251-255 bp).
# colnames(seqtab) are the DNA sequences; nchar() gets their lengths.
# %in% seq(251, 255) checks if each length is 251, 252, 253, 254, or 255.
# This removes unusually short or long sequences that are likely artifacts.
seqtab <- seqtab[, nchar(colnames(seqtab)) %in% seq(251, 255)]
cat("ASVs remaining after length filtering:", ncol(seqtab), "\n")
```

### Chimera removal

**Chimeras** are hybrid sequences formed during PCR when an incomplete extension
product re-anneals to a different template in a subsequent cycle. They can
constitute ~30% of unique sequences in raw data.

`removeBimeraDenovo()` identifies chimeras by checking whether each ASV can be
reconstructed as a combination of two more abundant "parent" sequences. Because
chimeric sequences are individually low-abundance, the total read loss is
usually modest.

```{r chimera-removal}
# removeBimeraDenovo() identifies and removes chimeric sequences.
# Chimeras are artificial sequences created during PCR when an incomplete DNA
# copy from one template accidentally joins with a different template. The result
# is a hybrid (chimera) of two real sequences.
# method = "consensus" checks each ASV across all samples — if an ASV is
# flagged as chimeric in a majority of samples where it appears, it is removed.
rds_seqtab_nochim <- file.path(dada2_rds_dir, "08_seqtab_nochim.rds")
if (use_precomputed_dada2) {
  seqtab.nochim <- require_rds_checkpoint(rds_seqtab_nochim, "chimera-filtered sequence table")
  message("Loaded precomputed chimera-filtered table: ", rds_seqtab_nochim)
} else if (file.exists(rds_seqtab_nochim)) {
  seqtab.nochim <- readRDS(rds_seqtab_nochim)
  message("Loaded precomputed chimera-filtered table: ", rds_seqtab_nochim)
} else {
  seqtab.nochim <- dada2::removeBimeraDenovo(
    seqtab,                 # Input ASV table (before chimera removal)
    method    = "consensus", # Use consensus across samples to flag chimeras
    multithread = threads,   # Parallelize for speed
    verbose   = TRUE         # Print progress messages
  )
  saveRDS(seqtab.nochim, rds_seqtab_nochim)
}
# Optional fast restart (uncomment if this step is too slow):
# if (file.exists(file.path(dada2_rds_dir, "08_seqtab_nochim.rds"))) {
#   seqtab.nochim <- readRDS(file.path(dada2_rds_dir, "08_seqtab_nochim.rds"))
# }

# Report how many ASVs were identified as chimeras and removed.
# Many unique sequences may be chimeric, but they are usually low-abundance,
# so the total number of READS lost is typically small (< 15%).
cat("\nASVs before chimera removal:", ncol(seqtab), "\n")
cat("ASVs after chimera removal:", ncol(seqtab.nochim), "\n")
cat("Fraction of reads retained:",
    round(sum(seqtab.nochim) / sum(seqtab) * 100, 1), "%\n")
```

### Track reads through the pipeline

Building a tracking table lets us verify that a reasonable fraction of reads
survived each step. Expect **>60% overall pass-through**. Large drops at
specific steps indicate problems:

- Large loss at filtering → overly aggressive quality thresholds
- Large loss at merging → insufficient overlap (over-trimming) or poor R2 quality
- Large loss at chimera removal → potential contamination or PCR issues

```{r tracking-table}
# Helper function: extracts the total number of reads from a dada or merger object
getN <- function(x) sum(dada2::getUniques(x))

# Build a summary table showing how many reads survived each pipeline step.
# In precomputed mode, align samples using run_accession <-> sample_id mapping.
track_samples <- Reduce(intersect, list(names(dadaFs), names(mergers), rownames(seqtab), rownames(seqtab.nochim)))
if (length(track_samples) == 0) {
  stop("No overlapping sample IDs found across DADA2 objects for tracking.")
}

out_df <- as.data.frame(out, stringsAsFactors = FALSE)
if (ncol(out_df) < 2 || is.null(rownames(out_df))) {
  stop("filterAndTrim checkpoint does not have expected row/column structure.")
}
in_col <- colnames(out_df)[1]
out_col <- colnames(out_df)[2]
out_df$run_accession <- sub("_[12]\\.fastq\\.gz$", "", rownames(out_df))
sample_map <- samplesheet_map %>%
  dplyr::transmute(
    run_accession = as.character(.data$run_accession),
    sample_id = normalize_sample_id(.data$sample_id)
  ) %>%
  dplyr::distinct(.data$run_accession, .keep_all = TRUE)
out_df <- out_df %>%
  dplyr::left_join(sample_map, by = "run_accession")
out_summary <- out_df %>%
  dplyr::filter(.data$sample_id %in% track_samples) %>%
  dplyr::group_by(.data$sample_id) %>%
  dplyr::summarise(
    input = as.numeric(max(.data[[in_col]], na.rm = TRUE)),
    filtered = as.numeric(max(.data[[out_col]], na.rm = TRUE)),
    .groups = "drop"
  )

if (nrow(out_summary) != length(track_samples)) {
  missing_track <- setdiff(track_samples, out_summary$sample_id)
  stop(
    "Failed to map filterAndTrim counts to sample IDs for tracking. Missing: ",
    paste(utils::head(missing_track, 8), collapse = ", ")
  )
}

track <- data.frame(
  sample_id = track_samples,
  denoised = sapply(dadaFs, getN)[track_samples],    # Reads after denoising
  merged   = sapply(mergers, getN)[track_samples],   # Reads after merging
  tabled   = rowSums(seqtab)[track_samples],         # Reads in sequence table
  nochim   = rowSums(seqtab.nochim)[track_samples]   # Reads after chimera removal
) %>%
  dplyr::left_join(out_summary, by = "sample_id") %>%
  dplyr::select(.data$sample_id, .data$input, .data$filtered, .data$denoised, .data$merged, .data$tabled, .data$nochim)
rownames(track) <- track$sample_id
track$sample_id <- NULL

# Calculate what percentage of the original reads made it through the entire pipeline
track$percent_retained <- round(track$nochim / track$input * 100, 1)
utils::write.csv(track, file.path(dada2_tbl_dir, "02_read_tracking_table.csv"), row.names = TRUE)

# Visualize read retention by sample across pipeline stages.
track$sample_id <- rownames(track)
track_long <- track %>%
  tibble::as_tibble() %>%
  tidyr::pivot_longer(
    cols = c("input", "filtered", "denoised", "merged", "tabled", "nochim"),
    names_to = "stage",
    values_to = "reads"
  ) %>%
  dplyr::mutate(
    stage = factor(
      .data$stage,
      levels = c("input", "filtered", "denoised", "merged", "tabled", "nochim"),
      labels = c("Input", "Filtered", "Denoised", "Merged", "Seq table", "No chimera")
    )
  )

stage_summary <- track_long %>%
  dplyr::group_by(.data$stage) %>%
  dplyr::summarise(median_reads = stats::median(.data$reads), .groups = "drop")

ggplot2::ggplot(track_long, ggplot2::aes(x = .data$stage, y = .data$reads, group = .data$sample_id)) +
  ggplot2::geom_line(alpha = 0.25, color = "#2C7FB8") +
  ggplot2::geom_point(alpha = 0.35, size = 1.4, color = "#2C7FB8") +
  ggplot2::geom_line(
    data = stage_summary,
    ggplot2::aes(x = .data$stage, y = .data$median_reads, group = 1),
    inherit.aes = FALSE,
    color = "#D95F0E",
    linewidth = 1.2
  ) +
  ggplot2::geom_point(
    data = stage_summary,
    ggplot2::aes(x = .data$stage, y = .data$median_reads),
    inherit.aes = FALSE,
    color = "#D95F0E",
    size = 2.2
  ) +
  ggplot2::scale_y_continuous(labels = scales::comma) +
  ggplot2::theme_minimal() +
  ggplot2::labs(
    title = "Read Retention Across DADA2 Stages",
    subtitle = "Blue lines = individual samples; orange line = median across samples",
    x = "Pipeline stage",
    y = "Read count"
  )

# Display the tracking table — expect >60% retention for a healthy dataset
track
```

**Discussion prompts for students**

- At which stage do most samples lose the largest fraction of reads, and why might that be expected biologically or technically?
- Do any samples show a much steeper drop than the others at a specific stage? What could that indicate about those samples?
- If read loss is high at the merging step, what parameter choices or read-quality issues would you inspect first?
- How would your interpretation change if controls (blanks/mock) lose reads differently than biological samples?

### Assign taxonomy

Taxonomy is assigned using a **naive Bayesian classifier** (Wang et al., 2007)
that compares each ASV sequence against the SILVA v138 reference database.
`addSpecies()` attempts exact matching for species-level assignments.

Some ASVs will lack genus- or species-level classification---this is normal.
Short 16S fragments (especially from V4 alone) cannot always resolve to species,
and some sequences simply lack a close match in the reference database.

```{r assign-taxonomy}
# assignTaxonomy() takes each ASV sequence and compares it against a reference
# database (SILVA) to determine its taxonomy (Kingdom > Phylum > Class > ... > Genus).
# It uses a naive Bayesian classifier: essentially asking "given this sequence,
# what is the probability it belongs to each known taxon?"
#
# assignTaxonomy() takes ~5 minutes.
silva_taxonomy_db <- file.path(data_base_dir, "Databases", "silva_nr99_v138.2_toSpecies_trainset.fa.gz")
if (!file.exists(silva_taxonomy_db)) {
  stop("SILVA taxonomy database not found at: ", silva_taxonomy_db)
}
rds_taxa <- file.path(dada2_rds_dir, "09_taxa.rds")
if (use_precomputed_dada2) {
  taxa <- require_rds_checkpoint(rds_taxa, "taxonomy assignments")
  message("Loaded precomputed taxonomy checkpoint: ", rds_taxa)
} else if (file.exists(rds_taxa)) {
  taxa <- readRDS(rds_taxa)
  message("Loaded precomputed taxonomy checkpoint: ", rds_taxa)
} else {
  set.seed(42)  # For reproducibility of the classifier's bootstrap sampling
  taxa <- dada2::assignTaxonomy(
    seqtab.nochim,                                 # Our chimera-free ASV table
    silva_taxonomy_db,                              # SILVA v138.2 toSpecies training set
    multithread = threads                           # Parallelize for speed
  )

  # Optional exact species matching (only run if the species assignment file exists).
  silva_species_db <- file.path(data_base_dir, "Databases", "silva_species_assignment_v138.2.fa.gz")
  if (file.exists(silva_species_db)) {
    taxa <- dada2::addSpecies(taxa, silva_species_db)
  }
  saveRDS(taxa, rds_taxa)
}
# Optional fast restart (uncomment if this step is too slow):
# if (file.exists(file.path(dada2_rds_dir, "09_taxa.rds"))) {
#   taxa <- readRDS(file.path(dada2_rds_dir, "09_taxa.rds"))
# }

# Preview the taxonomy table: each row is an ASV, columns are taxonomic ranks
# (Kingdom, Phylum, Class, Order, Family, Genus, Species)
# NA means the classifier couldn't confidently assign that rank
taxa.print <- taxa
rownames(taxa.print) <- NULL  # Remove long sequence names for cleaner display
head(taxa.print)
```

------------------------------------------------------------------------

## Filtering and Building the Phyloseq Object

### Create a preliminary phyloseq object

The `phyloseq` package provides a unified data structure that bundles the ASV
table, taxonomy, sample metadata, and optionally a phylogenetic tree into a
single R object. Most downstream microbiome analysis tools accept phyloseq
objects directly.

```{r build-initial-phyloseq}
# For Part 1, continue with the objects generated above.
seqtab.nochim.full <- seqtab.nochim
taxa.full <- taxa
# Optional fast restart (uncomment if checkpoints exist):
# seqtab.nochim.full <- readRDS(file.path(dada2_rds_dir, "08_seqtab_nochim.rds"))
# taxa.full <- readRDS(file.path(dada2_rds_dir, "09_taxa.rds"))

# phyloseq() creates a unified data object that links together:
#   - otu_table: the ASV count matrix (samples x ASVs)
#   - tax_table: the taxonomy for each ASV (Kingdom through Species)
# taxa_are_rows = FALSE tells phyloseq that SAMPLES are rows and ASVs are columns
# (this matches the DADA2 output format)
ps <- phyloseq::phyloseq(
  phyloseq::otu_table(seqtab.nochim.full, taxa_are_rows = FALSE),
  phyloseq::tax_table(taxa.full)
)
# Ensure sample names remain non-syntactic-safe IDs (e.g., keep "01" not "X01").
phyloseq::sample_names(ps) <- normalize_sample_id(phyloseq::sample_names(ps))

# By default, ASV names are the full DNA sequences (e.g., "TACGGAGG..."),
# which are unwieldy. We replace them with short IDs like "ASV1", "ASV2", etc.
# We save the mapping between sequence and ID in case we need it later.
names_df <- data.frame(
  seq = phyloseq::taxa_names(ps),                         # Original: full DNA sequences
  asv = paste0("ASV", seq_len(phyloseq::ntaxa(ps))),      # New: ASV1, ASV2, ...
  stringsAsFactors = FALSE
)
phyloseq::taxa_names(ps) <- names_df$asv  # Apply the new short names
saveRDS(ps, file.path(dada2_rds_dir, "10_ps_initial.rds"))
saveRDS(names_df, file.path(dada2_rds_dir, "11_asv_name_map.rds"))
# Optional fast restart (uncomment if this step is too slow):
# ps <- readRDS(file.path(dada2_rds_dir, "10_ps_initial.rds"))
# names_df <- readRDS(file.path(dada2_rds_dir, "11_asv_name_map.rds"))

# Summarize: how many ASVs and samples are in our initial phyloseq object?
cat("Initial phyloseq:", phyloseq::ntaxa(ps), "ASVs across",
    phyloseq::nsamples(ps), "samples\n")
```

### Remove low-depth samples and tag controls

```{r remove-low-depth}
# Attach metadata before QC so controls can be used in decontam/diagnostics.
meta_qc <- read.csv(metadata_path, stringsAsFactors = FALSE, check.names = FALSE)
stopifnot(all(c("Sample_ID", "Treatment", "Negative") %in% colnames(meta_qc)))
meta_qc$Sample_ID <- normalize_sample_id(meta_qc$Sample_ID)

# Align metadata rows to the phyloseq sample order.
sample_ids_ps <- normalize_sample_id(phyloseq::sample_names(ps))
missing_ids <- setdiff(sample_ids_ps, meta_qc$Sample_ID)
if (length(missing_ids) > 0) {
  stop("Metadata is missing Sample_ID values for: ", paste(missing_ids, collapse = ", "))
}
meta_qc <- meta_qc[match(sample_ids_ps, meta_qc$Sample_ID), ]
rownames(meta_qc) <- meta_qc$Sample_ID

# Derive explicit control flags for this workshop dataset:
# - kit blanks: N1..N9 / Treatment == "blank"
# - water blanks: H2O controls / Treatment == "H2O"
# - mock controls: MOCK_* / Treatment == "mock"
# - inoculum controls: Treatment == "inoculum"
meta_qc$is_kit_blank <- grepl("^N[0-9]+$", meta_qc$Sample_ID) | meta_qc$Treatment == "blank"
meta_qc$is_water_blank <- grepl("^H2O", meta_qc$Sample_ID) | meta_qc$Treatment == "H2O"
meta_qc$is_mock_control <- grepl("^MOCK", meta_qc$Sample_ID) | tolower(meta_qc$Treatment) == "mock"
meta_qc$is_inoculum_control <- tolower(meta_qc$Treatment) == "inoculum"
meta_qc$is_negative <- meta_qc$Negative %in% c(TRUE, "TRUE", "True", "true", 1, "1") |
  meta_qc$is_kit_blank | meta_qc$is_water_blank
meta_qc$is_control_any <- meta_qc$is_negative | meta_qc$is_mock_control | meta_qc$is_inoculum_control

# Add metadata and control flags to the phyloseq object.
phyloseq::sample_data(ps) <- phyloseq::sample_data(meta_qc)

# sample_sums() returns the total number of reads in each sample.
# Plotting this distribution helps choose a sensible minimum depth threshold.
read_depth <- phyloseq::sample_sums(ps)
depth_df <- data.frame(
  sample_id = names(read_depth),
  read_depth = as.numeric(read_depth),
  is_control = as.logical(phyloseq::sample_data(ps)$is_control_any)
)

# Histogram of read depth across all samples.
# Biological and control samples are shown together so users can see the full range.
ggplot2::ggplot(depth_df, ggplot2::aes(x = read_depth)) +
  ggplot2::geom_histogram(
    bins = 30,
    fill = RColorBrewer::brewer.pal(8, "Dark2")[3],
    color = "white"
  ) +
  ggplot2::theme_minimal() +
  ggplot2::labs(
    title = "Read Depth Distribution Across Samples",
    x = "Total reads per sample",
    y = "Number of samples"
  )

# Sort depths so users can inspect specific low-depth samples.
sort(read_depth)

# Default minimum depth threshold for biological samples.
# Adjust this value based on the histogram and project goals.
min_read_depth <- 5000
cat("Using default minimum read depth:", min_read_depth, "\n")

# prune_samples() removes low-depth biological samples.
# Controls are retained regardless of depth so they remain available for decontam/QC.
ps.pruned <- ps %>%
  phyloseq::prune_samples(
    phyloseq::sample_sums(ps) > min_read_depth | phyloseq::sample_data(ps)$is_control_any, .
  )

cat("Samples after depth filter:", phyloseq::nsamples(ps.pruned), "\n")
cat("Negative controls kept for decontam:", sum(phyloseq::sample_data(ps.pruned)$is_negative), "\n")
cat("Mock controls kept for QC:", sum(phyloseq::sample_data(ps.pruned)$is_mock_control), "\n")
saveRDS(ps.pruned, file.path(dada2_rds_dir, "12_ps_pruned.rds"))
# Optional fast restart (uncomment if this step is too slow):
# ps.pruned <- readRDS(file.path(dada2_rds_dir, "12_ps_pruned.rds"))
```

### Contaminant identification with decontam

Laboratory reagents, DNA extraction kits, and the sequencing environment itself
can introduce microbial DNA that is not part of your biological samples. The
`decontam` package (Davis et al., 2018) identifies likely contaminant ASVs using
two complementary strategies:

| Method | Requires | Logic |
|--------|----------|-------|
| **Frequency** | DNA concentration (e.g., PicoGreen) | Contaminants are *more* abundant in low-concentration samples (inverse correlation with DNA input) |
| **Prevalence** | Negative control samples | Contaminants are *more* prevalent in negative controls than in true samples |

You can run either method alone, or combine both with `method = "either"` for
maximum sensitivity. If you have both DNA concentration data and negative
controls, combining is recommended.

> **Note:** This dataset includes kit blanks, water blanks, and mock controls in
> metadata, so the prevalence/frequency decontam workflow is applicable. The
> chunk remains `eval=FALSE` in the tutorial to keep runtime manageable.

```{r decontam-example, eval=FALSE}
# --- Frequency-based method ---
# Logic: contaminant DNA comes from reagents at a roughly fixed amount,
# so in samples with LESS total DNA (low concentration), contaminant sequences
# make up a LARGER fraction. isContaminant() looks for this inverse correlation
# between DNA concentration and sequence abundance.
contamdf_freq <- decontam::isContaminant(
  ps.pruned,              # Phyloseq object to test
  method = "frequency",   # Use the frequency-based approach
  conc   = "Picogreen"   # Name of the column in sample_data() holding DNA concentrations
)
# table() counts how many ASVs are flagged as contaminants (TRUE) vs. not (FALSE)
table(contamdf_freq$contaminant)

# --- Prevalence-based method ---
# Logic: contaminant sequences should appear MORE OFTEN in negative control
# samples (which contain no biological DNA) than in real samples.
contamdf_prev <- decontam::isContaminant(
  ps.pruned,
  method    = "prevalence",    # Use the prevalence-based approach
  neg       = "is_negative",  # Column in sample_data(): TRUE = negative control sample
  normalize = TRUE,            # Normalize for differences in sequencing depth
  threshold = 0.1             # P-value threshold; lower = stricter (fewer contaminants flagged)
)
table(contamdf_prev$contaminant)

# Visualize which ASVs are contaminants by plotting their prevalence in
# real samples (y-axis) vs. negative controls (x-axis).
# Contaminants should cluster toward the upper-left (common in negatives, rare in real).

# transform_sample_counts() converts counts to presence/absence (1 or 0)
ps_pa     <- phyloseq::transform_sample_counts(ps.pruned, function(x) 1 * (x > 0))
# Split into negative controls and real (positive) samples
ps_pa_neg <- phyloseq::prune_samples(sample_data(ps_pa)$is_negative == TRUE, ps_pa)
ps_pa_pos <- phyloseq::prune_samples(sample_data(ps_pa)$is_negative == FALSE, ps_pa)

# taxa_sums() on presence/absence data = number of samples each ASV appears in
prev_df <- data.frame(
  pa_pos      = phyloseq::taxa_sums(ps_pa_pos),    # Prevalence in real samples
  pa_neg      = phyloseq::taxa_sums(ps_pa_neg),    # Prevalence in negative controls
  contaminant = contamdf_prev$contaminant           # TRUE/FALSE contaminant flag
)
ggplot2::ggplot(prev_df, ggplot2::aes(x = pa_neg, y = pa_pos, color = contaminant)) +
  ggplot2::geom_point(size = 2, alpha = 0.7) +
  ggplot2::scale_color_brewer(palette = "Dark2") +
  ggplot2::labs(x = "Prevalence (Negative Controls)",
                y = "Prevalence (True Samples)") +
  ggplot2::theme_minimal()

# --- Combined method (recommended when both data types are available) ---
# method = "either" flags an ASV as a contaminant if EITHER the frequency
# or prevalence method identifies it, giving maximum sensitivity.
contamdf_both <- decontam::isContaminant(
  ps.pruned,
  conc      = "Picogreen",    # DNA concentration column
  neg       = "is_negative",  # Negative control indicator column
  method    = "either",       # Flag if EITHER method flags it
  normalize = TRUE
)
table(contamdf_both$contaminant)

# Remove all ASVs identified as contaminants from the phyloseq object.
# The "!" negates the logical vector, keeping only non-contaminant ASVs.
ps.decontam <- phyloseq::prune_taxa(!contamdf_both$contaminant, ps.pruned)
cat("ASVs removed by decontam:", sum(contamdf_both$contaminant), "\n")
cat("ASVs remaining:", phyloseq::ntaxa(ps.decontam), "\n")
```

> **Why decontam matters:** Contaminant sequences can dominate low-biomass
> samples, create spurious associations in differential abundance analysis, and
> inflate alpha diversity. Even in high-biomass samples, contaminants add noise.
> Running decontam is especially critical for:
>
> - Environmental samples with low microbial biomass
> - Studies using kit-based DNA extraction (the "kitome")
> - Any study where negative controls were sequenced

For workshop timing, we skip running decontam live and proceed to taxa
filtering. On a full run, execute the decontam chunk above and continue with the
decontaminated object.

### Optional: Mock community QC with plasmid references

If you have a reference FASTA for the mock community (for example,
`Data/Databases/mock_seq_plasmids.fasta`), you can use it as an additional QC
layer. This is not required for the main workflow, but it helps verify whether
dominant mock-sample ASVs are consistent with expected mock reference sequences.

```{r mock-community-qc-optional, eval=FALSE}
mock_fasta_path <- file.path(data_base_dir, "Databases", "mock_seq_plasmids.fasta")
if (!file.exists(mock_fasta_path)) {
  stop("Mock reference FASTA not found at: ", mock_fasta_path)
}

if (!exists("meta_qc")) {
  stop("meta_qc not found. Run the remove-low-depth chunk first.")
}

# Official DADA2-style reference loading for exact-sequence comparison.
mock_ref <- dada2::getSequences(mock_fasta_path)

# Identify mock samples from metadata and keep those that are present in seqtab.nochim.
mock_samples <- intersect(
  rownames(meta_qc)[meta_qc$is_mock_control],
  rownames(seqtab.nochim)
)
if (length(mock_samples) == 0) {
  stop("No mock samples detected in seqtab.nochim.")
}

# Pool mock counts across all detected mock-control samples.
unqs_mock <- colSums(seqtab.nochim[mock_samples, , drop = FALSE])
unqs_mock <- sort(unqs_mock[unqs_mock > 0], decreasing = TRUE)  # Drop ASVs absent in mock samples

cat("DADA2 inferred", length(unqs_mock), "sequence variants present in mock sample(s).\n")

# Official DADA2-style exact-match evaluation.
match_ref <- sum(sapply(names(unqs_mock), function(x) any(x == mock_ref)))
cat("Of those,", match_ref, "were exact matches to expected mock reference sequences.\n")

# Export the core mock-accuracy summary.
mock_eval_tbl <- data.frame(
  n_mock_samples = length(mock_samples),
  n_inferred_mock_asvs = length(unqs_mock),
  n_exact_reference_matches = match_ref,
  frac_exact_reference_matches = round(match_ref / length(unqs_mock), 4),
  stringsAsFactors = FALSE
)
utils::write.csv(
  mock_eval_tbl,
  file.path(dada2_tbl_dir, "04_mock_accuracy_summary.csv"),
  row.names = FALSE
)

# Export detailed ASV-level table with taxonomy and exact-match flag.
mock_asv_tbl <- data.frame(
  asv_sequence = names(unqs_mock),
  mock_abundance = as.numeric(unqs_mock),
  exact_reference_match = names(unqs_mock) %in% mock_ref,
  stringsAsFactors = FALSE
) %>%
  dplyr::left_join(
    as.data.frame(taxa) %>%
      tibble::rownames_to_column("asv_sequence"),
    by = "asv_sequence"
  ) %>%
  dplyr::arrange(dplyr::desc(.data$mock_abundance))

utils::write.csv(
  mock_asv_tbl,
  file.path(dada2_tbl_dir, "05_mock_asvs_with_taxonomy_and_match_flag.csv"),
  row.names = FALSE
)

head(mock_asv_tbl)
```

### Post-DADA2 taxa filtering

After building the initial phyloseq object (and optionally running decontam),
we apply a series of quality filters to remove sequences that are not
informative for microbial community analysis. Each step targets a specific
category of problematic ASV.

We report the number of ASVs removed at each step for transparency---this is
good practice for methods sections and supplementary materials.

```{r taxa-filtering-pipeline}
# Record how many ASVs we start with so we can track how many are removed
n_start <- phyloseq::ntaxa(ps.pruned)
cat(sprintf("Starting ASVs: %d\n\n", n_start))

# --- Step 1: Prevalence filter ---
# Remove ASVs found in fewer than N samples.
# Very rare ASVs (present in only 1-2 samples) are often:
#   - Sequencing artifacts that escaped DADA2's error model
#   - Contaminants from reagents or the environment
#   - Transient environmental DNA, not true community members
# A good threshold is the minimum group size in your experimental design.
prev_threshold <- 3  # ASV must appear in at least 3 samples to be kept
n_before <- phyloseq::ntaxa(ps.pruned)

# apply() loops over the ASV table to count in how many samples each ASV appears.
# MARGIN = 1 means loop over rows, MARGIN = 2 means columns.
# taxa_are_rows() tells us which dimension is ASVs so we loop over the right one.
# sum(x > 0) counts how many samples have at least 1 read for that ASV.
prevalence <- apply(
  X = phyloseq::otu_table(ps.pruned),
  MARGIN = ifelse(phyloseq::taxa_are_rows(ps.pruned), yes = 1, no = 2),
  FUN = function(x) sum(x > 0)
)

# Keep only ASVs that pass the prevalence threshold
keep_taxa <- names(prevalence[prevalence >= prev_threshold])
ps.filt <- phyloseq::prune_taxa(keep_taxa, ps.pruned)  # Subset the phyloseq object
cat(sprintf("Step 1 - Prevalence filter (min %d samples): %d -> %d ASVs (%d removed)\n",
            prev_threshold, n_before, phyloseq::ntaxa(ps.filt),
            n_before - phyloseq::ntaxa(ps.filt)))

# --- Step 2: Remove non-target taxa (Mitochondria, Chloroplast, Eukaryota) ---
# 16S primers can accidentally amplify DNA from:
#   - Host mitochondria (mitochondria evolved from bacteria, so they have 16S genes)
#   - Plant/algal chloroplasts (also of bacterial origin)
#   - Eukaryotes (cross-amplification of 18S rRNA genes)
# These are not part of the microbial community we are studying and must be
# removed to avoid inflating diversity estimates and confounding comparisons.
n_before <- phyloseq::ntaxa(ps.filt)

# subset_taxa() filters ASVs based on their taxonomy table values.
# We keep only Bacteria and Archaea, and exclude Mitochondria and Chloroplast.
# We use %in% (not ==) so that ASVs with NA taxonomy are not accidentally removed.
ps.filt <- ps.filt %>%
  phyloseq::subset_taxa(
    Kingdom %in% c("Bacteria", "Archaea") &  # Only keep bacterial/archaeal sequences
    !Family %in% "Mitochondria" &             # Remove mitochondrial sequences
    !Order %in% "Chloroplast"                 # Remove chloroplast sequences
  )
cat(sprintf("Step 2 - Remove Mitochondria/Chloroplast/non-Bacteria: %d -> %d ASVs (%d removed)\n",
            n_before, phyloseq::ntaxa(ps.filt),
            n_before - phyloseq::ntaxa(ps.filt)))

# --- Step 3: Remove ASVs with no phylum-level classification ---
# ASVs classified only to Kingdom (e.g., "Bacteria") with no Phylum assignment
# are typically:
#   - Highly divergent sequences with no close database match
#   - Chimeras or artifacts that passed the earlier filter
#   - Sequences too short or degraded for reliable classification
# These provide no meaningful taxonomic information and add noise to downstream
# analyses that group results by phylum or lower taxonomic ranks.
n_before <- phyloseq::ntaxa(ps.filt)
ps.filt <- ps.filt %>%
  phyloseq::subset_taxa(!is.na(Phylum) & Phylum != "")  # Keep only ASVs with a Phylum assignment
cat(sprintf("Step 3 - Remove ASVs with no Phylum classification: %d -> %d ASVs (%d removed)\n",
            n_before, phyloseq::ntaxa(ps.filt),
            n_before - phyloseq::ntaxa(ps.filt)))

# --- Summary ---
# Report the overall effect of all filtering steps combined
n_end <- phyloseq::ntaxa(ps.filt)
cat(sprintf("\nFiltering complete: %d -> %d ASVs (removed %d total, %.1f%%)\n",
            n_start, n_end, n_start - n_end,
            100 * (n_start - n_end) / n_start))
saveRDS(ps.filt, file.path(dada2_rds_dir, "13_ps_filt_pre_metadata.rds"))
# Optional fast restart (uncomment if this step is too slow):
# ps.filt <- readRDS(file.path(dada2_rds_dir, "13_ps_filt_pre_metadata.rds"))
```

> **Why each filter matters:**
>
> | Filter | Target | Risk if skipped |
> |--------|--------|-----------------|
> | Prevalence | Singletons, rare artifacts | Inflated richness, spurious DA hits |
> | Mitochondria/Chloroplast | Host organellar DNA | Inflated diversity, wrong kingdom |
> | No Phylum | Unclassifiable sequences | Noise in taxonomic summaries, uninterpretable results |
>
> Always report the number of ASVs removed at each step in your methods section.
> Large removals at any step may indicate upstream problems (e.g., many
> Mitochondria suggests host contamination; many no-Phylum suggests poor
> sequence quality or an inappropriate reference database).

### Add sample metadata and finalize

```{r add-metadata}
# Read metadata table and align it to the DADA2 sample IDs from samplesheet.
metadata <- read.csv(metadata_path, stringsAsFactors = FALSE, check.names = FALSE)
stopifnot(all(c("Sample_ID", "Treatment", "Negative") %in% colnames(metadata)))
metadata$Sample_ID <- normalize_sample_id(metadata$Sample_ID)

# Match metadata rows to the exact sample order currently present in ps.filt.
sample_ids_ps <- normalize_sample_id(phyloseq::sample_names(ps.filt))
missing_ids <- setdiff(sample_ids_ps, metadata$Sample_ID)
if (length(missing_ids) > 0) {
  stop("Metadata is missing Sample_ID values for: ", paste(missing_ids, collapse = ", "))
}
metadata <- metadata[match(sample_ids_ps, metadata$Sample_ID), ]
rownames(metadata) <- metadata$Sample_ID

# Recreate control flags in case this chunk is run independently.
metadata$is_kit_blank <- grepl("^N[0-9]+$", metadata$Sample_ID) | metadata$Treatment == "blank"
metadata$is_water_blank <- grepl("^H2O", metadata$Sample_ID) | metadata$Treatment == "H2O"
metadata$is_mock_control <- grepl("^MOCK", metadata$Sample_ID) | tolower(metadata$Treatment) == "mock"
metadata$is_inoculum_control <- tolower(metadata$Treatment) == "inoculum"
metadata$is_negative <- metadata$Negative %in% c(TRUE, "TRUE", "True", "true", 1, "1") |
  metadata$is_kit_blank | metadata$is_water_blank
metadata$is_control_any <- metadata$is_negative | metadata$is_mock_control | metadata$is_inoculum_control
metadata$Treatment <- as.factor(metadata$Treatment)
if ("Hive" %in% colnames(metadata)) metadata$Hive <- as.factor(metadata$Hive)
if ("Box" %in% colnames(metadata)) metadata$Box <- as.factor(metadata$Box)

# Attach metadata to the phyloseq object.
phyloseq::sample_data(ps.filt) <- phyloseq::sample_data(metadata)

# Remove control samples before downstream ecological comparisons.
sample_meta_df <- data.frame(phyloseq::sample_data(ps.filt), check.names = FALSE)
control_flags <- sample_meta_df$is_control_any
ps.filt <- phyloseq::prune_samples(!control_flags, ps.filt)
cat("Samples retained for biological analyses:", phyloseq::nsamples(ps.filt), "\n")
saveRDS(ps.filt, file.path(dada2_rds_dir, "14_ps_filt_final.rds"))
# Optional fast restart (uncomment if this step is too slow):
# ps.filt <- readRDS(file.path(dada2_rds_dir, "14_ps_filt_final.rds"))
sample_meta_df <- data.frame(phyloseq::sample_data(ps.filt), check.names = FALSE)
utils::write.csv(
  sample_meta_df,
  file.path(dada2_tbl_dir, "03_sample_metadata_after_control_filter.csv"),
  row.names = TRUE
)

# Explore the final phyloseq object — it now contains three linked components:
# 1) The ASV count table (otu_table)
# 2) The taxonomy assignments (tax_table)
# 3) The sample metadata (sample_data)
ps.filt
cat("\n--- OTU table preview (first 5 samples x first 5 ASVs) ---\n")
phyloseq::otu_table(ps.filt)[1:5, 1:5]
cat("\n--- Taxonomy table preview (first 6 ASVs) ---\n")
head(phyloseq::tax_table(ps.filt))
cat("\n--- Sample data preview (first 6 samples) ---\n")
head(phyloseq::sample_data(ps.filt))
```

> **Congratulations!** You now have a fully processed phyloseq object ready for
> ecological analysis. This is the end product of Part 1. In a real analysis,
> you would save this object with `saveRDS(ps.filt, "my_phyloseq.rds")`.

------------------------------------------------------------------------

# PART 2: From Phyloseq to Biological Insights {.tabset}

```{r set-part2-figure-path, include=FALSE}
# Route Part 2 figures to Results/MicrobiomeAnalysis/Figures
knitr::opts_chunk$set(fig.path = file.path(microbiome_fig_dir, "fig-"))
```

We now shift from data processing to **biological analysis**. Starting from a
phyloseq object, we will ask three progressively specific questions about our
microbial communities:

1. **Alpha diversity:** How diverse is each sample, and do groups differ?
2. **Beta diversity:** Do overall community structures differ between groups?
3. **Differential abundance:** Which specific taxa drive those differences?

### Setup for Part 2

There are many valid ways to explore and analyze microbiome data, and your
choice of tools should match your biological question and study design. A common
foundation in R is the **phyloseq** package. In Part 1, we used DADA2 outputs
(ASV/OTU table, taxonomy table, and sample metadata) to build a phyloseq object,
which serves as a unified data structure for downstream analysis.

Many newer workflows leverage phyloseq objects directly, including packages such
as **microViz**, **microbiome**, **vegan** and methods that integrate through phyloseq
objects (for example **DESeq2** and **ANCOM-BC** workflows). In this tutorial,
we emphasize **microViz** because it is user friendly and acts as a practical
wrapper around common microbiome analysis and visualization steps.

```{r load-libraries-part2}
library(phyloseq)     # Microbiome data structure and manipulation
library(vegan)        # Ecological statistics: PERMANOVA, beta-dispersion, ordination
library(ggplot2)      # Plotting and data visualization
library(RColorBrewer) # Color palettes (Dark2 prioritized for discrete groups)
library(tidyverse)    # Data wrangling (dplyr, tidyr, etc.)
library(data.table)   # Fast data manipulation (used in the differential abundance section)
library(microViz)     # Helper tools for working with phyloseq metadata and microbiome visual summaries

# Load a pre-built phyloseq object for Part 2.
# By default, we reuse ps.filt from Part 1 to keep sample IDs and metadata aligned.
if (exists("ps.filt")) {
  bee_ps <- ps.filt
} else {
  stop("`ps.filt` not found. Run Part 1 first (through 'add-metadata').")
}
# Optional fast restart (uncomment if this step is too slow):
# bee_ps <- readRDS(file.path(microbiome_rds_dir, "01_bee_ps_start.rds"))

# Fix tax_table issues once at the start of Part 2 so downstream microViz tools
# (e.g., tax_agg / tax_top / comp_barplot) do not fail on missing/uninformative taxa.
bee_ps <- bee_ps %>%
  microViz::tax_fix()

# Ensure categorical variables are stored as factors rather than plain text.
# R uses factors for grouping in statistical tests and for controlling the
# order of categories in plots.
bee_meta <- data.frame(phyloseq::sample_data(bee_ps), check.names = FALSE)
bee_meta$Treatment <- as.factor(bee_meta$Treatment)
if ("Hive" %in% colnames(bee_meta)) {
  bee_meta$Hive <- as.factor(bee_meta$Hive)
}
phyloseq::sample_data(bee_ps) <- phyloseq::sample_data(bee_meta)

# Print a summary of the phyloseq object: number of samples, ASVs, and metadata columns
bee_ps
saveRDS(bee_ps, file.path(microbiome_rds_dir, "01_bee_ps_start.rds"))
```

------------------------------------------------------------------------

## Data Exploration: Composition Overview

Before running any statistical tests, we should explore our data visually. This
helps identify obvious patterns, potential outliers, and sequencing depth issues.

### Sequencing depth

Uneven sequencing depth across samples can bias diversity estimates. Samples with
very few reads should be removed.

```{r sequencing-depth, fig.height=4}
# sample_sums() calculates the total number of reads per sample.
# Large differences in sequencing depth between samples can bias diversity
# estimates (samples with more reads tend to detect more taxa).
depth <- phyloseq::sample_sums(bee_ps)

# summary() gives min, max, median, and quartiles of the depth distribution
cat("Sequencing depth summary:\n")
summary(depth)

# Create a data frame for ggplot (ggplot2 requires data frames, not named vectors)
depth_df <- data.frame(
  sample = names(depth),        # Sample names
  depth  = as.numeric(depth)    # Read counts per sample
)

# Add treatment labels using an explicit sample-ID join (robust to unnamed vectors).
depth_meta <- data.frame(phyloseq::sample_data(bee_ps), check.names = FALSE) %>%
  tibble::rownames_to_column("sample") %>%
  dplyr::select(Treatment, sample)
depth_df <- depth_df %>%
  dplyr::left_join(depth_meta, by = "sample") %>%
  dplyr::arrange(Treatment, depth, sample) %>%
  dplyr::mutate(sample_order = factor(sample, levels = sample))

# Create a horizontal bar chart showing the depth of each sample, sorted from
# lowest to highest. This makes it easy to spot outlier samples with very few reads.
ggplot2::ggplot(depth_df, ggplot2::aes(x = depth, y = sample_order)) +
  ggplot2::geom_bar(
    stat = "identity",
    fill = RColorBrewer::brewer.pal(8, "Dark2")[2]
  ) +  # Bar height = read count
  # Optional exercise: uncomment to compare sequencing depth distributions by treatment.
  # ggplot2::facet_grid(. ~ Treatment, scales = "free_x") +
  ggplot2::coord_flip() +  # Flip axes so sample names are readable on the y-axis
  ggplot2::labs(x = "Sample", y = "Read count", title = "Sequencing Depth per Sample") +
  ggplot2::theme_minimal() +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 30, hjust = 1))

```

### Normalization overview

Raw count data cannot be directly compared across samples with different
sequencing depths. Common normalization strategies include:

| Method | Approach | Best for |
|--------|----------|----------|
| **Rarefaction** | Subsample to equal depth | Alpha/beta diversity |
| **Relative abundance** | Convert to proportions | Composition visualization |
| **CSS** | Cumulative sum scaling | General comparison |
| **DESeq2 VST** | Variance-stabilizing transform | Differential abundance |

For diversity analyses, we will use **rarefied data**. For differential
abundance, methods like DESeq2 handle normalization internally.

### Composition barplots

Stacked barplots provide a quick visual overview of which taxa dominate each
sample or group.

The first barplot shows **relative abundance**: each sample bar sums to 100%,
so it highlights proportional differences in community composition.

```{r composition-barplot-relative, fig.height=6, fig.width=10}
# microViz composition workflow:
# - aggregate to Genus
# - display top 8 genera + Other
bee_genus <- bee_ps %>%
  microViz::tax_agg(rank = "Genus")
saveRDS(bee_genus, file.path(microbiome_rds_dir, "03_bee_genus.rds"))
# Optional fast restart (uncomment if this step is too slow):
# bee_genus <- readRDS(file.path(microbiome_rds_dir, "03_bee_genus.rds"))

top_genera <- microViz::tax_top(bee_genus, n = 8, rank = "Genus")
genus_palette <- grDevices::colorRampPalette(RColorBrewer::brewer.pal(8, "Dark2"))(length(top_genera))
names(genus_palette) <- top_genera
genus_palette["Other"] <- "grey90"

bee_genus %>%
  microViz::comp_barplot(
    tax_level = "Genus",
    n_taxa = 8,
    other_name = "Other",
    merge_other = TRUE,
    palette = genus_palette,
    facet_by = "Treatment",
    sample_order = "asis",
    bar_outline_colour = "black"
  ) +
  ggplot2::theme_minimal() +
  ggplot2::theme(
    axis.text.x = ggplot2::element_text(angle = 30, hjust = 1, size = 9),  # Rotate sample labels
    legend.text = ggplot2::element_text(size = 8),                          # Smaller legend text
    legend.key.size = ggplot2::unit(0.4, "cm")
  ) +
  ggplot2::labs(
    x = "Sample",
    y = "Relative Abundance",
    title = "Relative Abundance of Top 8 Genera (+ Other)"
  )

```

> **For your own analyses, consider:** Compare the relative-abundance and absolute-abundance
> composition plots below. Which taxa appear similarly dominant in both views,
> and which change once total copy number is considered?

The next barplot shows an **absolute abundance proxy** at genus level by scaling
relative composition with per-sample total microbial load (`CopyNum_norm`).

```{r composition-barplot-absolute, fig.height=6, fig.width=10}

# Genus-level absolute abundance proxy:
# CopyNum_norm already stores sample-level absolute microbial load (normalized by gut cells).
# Here we distribute that total load across genera using each sample's genus composition.
# This does NOT re-estimate sample total abundance; it allocates the measured total to taxa.
copy_df <- data.frame(phyloseq::sample_data(bee_ps), check.names = FALSE) %>%
  tibble::rownames_to_column("Sample") %>%
  dplyr::mutate(
    total_copy = dplyr::coalesce(
      suppressWarnings(as.numeric(CopyNum_norm)),
      suppressWarnings(as.numeric(CopyNum))
    )
  ) %>% 
  dplyr::select(Sample, Treatment, total_copy)

if (all(is.na(copy_df$total_copy))) {
  warning("CopyNum / CopyNum_norm not found; skipping absolute abundance composition plot.")
} else {
  bee_genus_rel <- bee_genus %>%
    microViz::tax_transform(trans = "compositional")

  rel_genus_df <- phyloseq::psmelt(bee_genus_rel) %>%
    dplyr::mutate(
      Sample = as.character(Sample),
      Genus = dplyr::coalesce(as.character(Genus), "Unclassified"),
      Genus_plot = dplyr::if_else(Genus %in% top_genera, Genus, "Other")
    ) %>%
    dplyr::group_by(Sample, Treatment, Genus_plot) %>%
    dplyr::summarise(rel_abundance = sum(Abundance), .groups = "drop")

  comp_abs_df <- rel_genus_df %>%
    dplyr::left_join(copy_df %>% dplyr::select(Sample, total_copy), by = "Sample") %>%
    dplyr::mutate(
      abs_abundance = rel_abundance * total_copy,
      Sample = factor(Sample, levels = phyloseq::sample_names(bee_genus)),
      Genus_plot = factor(Genus_plot, levels = c(top_genera, "Other"))
    )

  p_abs <- ggplot2::ggplot(comp_abs_df, ggplot2::aes(x = Sample, y = abs_abundance, fill = Genus_plot)) +
    ggplot2::geom_col(width = 0.9, color = "black", linewidth = 0.1) +
    ggplot2::facet_wrap(~ Treatment, scales = "free_x", nrow = 1) +
    ggplot2::scale_fill_manual(
      values = genus_palette,
      breaks = c(top_genera, "Other"),
      name = "Genus"
    ) +
    ggplot2::scale_y_continuous(
      labels = scales::label_number(scale_cut = scales::cut_short_scale())
    ) +
    ggplot2::theme_minimal() +
    ggplot2::theme(
      axis.text.x = ggplot2::element_text(angle = 30, hjust = 1, size = 9),
      legend.text = ggplot2::element_text(size = 8),
      legend.key.size = ggplot2::unit(0.4, "cm")
    ) +
    ggplot2::labs(
      x = "Sample",
      y = "Genus-level abundance scaled by sample total load",
      title = "Top 8 Genera Scaled by Sample Absolute Load (+ Other)"
    )
  print(p_abs)
}
```

Even when relative-abundance profiles look similar, absolute-abundance profiles
can differ because samples may carry very different **total microbial loads**.
Relative abundance answers "what fraction of the community is this taxon?",
while absolute abundance answers "how many copies of this taxon are present?"
These are complementary, not interchangeable, views.

> **Interpreting composition plots:** These are descriptive, not statistical.
> Look for dominant taxa and obvious shifts between groups. Any patterns you
> observe must be confirmed with statistical tests below.
>
> **For your own analyses, consider:** Modify the composition workflow to use a
> different taxonomic rank (for example, `Phylum` or `Genus` instead of
> `Family`). How does the visual interpretation change when you zoom out
> (Phylum) versus zoom in (Genus/ASV)?
> [TODO: Add dataset-specific examples of expected dominant taxa if known.]

------------------------------------------------------------------------

## Alpha Diversity: Within-Sample Diversity

**Alpha diversity** measures the diversity *within* a single sample. Different
metrics capture different facets of community structure:

| Metric | Question | Sensitive to |
|--------|----------|-------------|
| **Observed** | "How many taxa are present?" | Rare taxa, sequencing depth |
| **Shannon (H')** | "How evenly distributed are taxa?" | Both richness and evenness |
| **Inverse Simpson (1/D)** | "How concentrated is abundance?" | Dominant taxa |

- **Observed richness** simply counts distinct ASVs. A sample with 50 ASVs has
  higher Observed than one with 30, regardless of their abundances.

- **Shannon index**: H' = -Σ(p~i~ × ln(p~i~)), where p~i~ is the proportion of
  species *i*. A community with 5 species at equal abundance
  (H' ≈ 1.6) has higher Shannon than one where a single species dominates
  (H' ≈ 1.2). Typical range for microbial communities: 0--5.

- **Inverse Simpson**: 1/D where D = Σ(p~i~^2^). Represents the probability
  that two randomly drawn individuals belong to different species. Higher values
  = more diverse. Less sensitive to rare taxa than Shannon.

### Calculate and visualize alpha diversity

This plot compares alpha diversity distributions between treatments for three
indices (Observed, Shannon, Inverse Simpson), with each point representing one sample.

```{r alpha-diversity, fig.height=5, fig.width=10}
# microViz workflow for alpha diversity:
# - do not filter taxa before calculating alpha diversity
# - calculate richness/diversity directly on phyloseq, then read from sample_data
alpha_ps <- bee_ps %>%
  microViz::ps_calc_richness(rank = "unique", index = "observed", varname = "Observed") %>%
  microViz::ps_calc_diversity(rank = "unique", index = "shannon", varname = "Shannon") %>%
  microViz::ps_calc_diversity(rank = "unique", index = "inverse_simpson", varname = "InvSimpson")

# Pull the calculated metrics + grouping variable from sample metadata.
alpha_div <- alpha_ps %>%
  microViz::samdat_tbl() %>%
  dplyr::transmute(
    sample = .sample_name,
    Treatment = as.character(Treatment),
    Observed = as.numeric(Observed),
    Shannon = as.numeric(Shannon),
    InvSimpson = as.numeric(InvSimpson)
  )

# Fail early if sample names do not align between richness output and metadata.
if (any(is.na(alpha_div$Treatment))) {
  stop(
    sprintf(
      "Treatment assignment failed for %d sample(s). Check sample IDs in bee_ps.",
      sum(is.na(alpha_div$Treatment))
    )
  )
}

# Reshape data from "wide" format (one column per metric) to "long" format
# (one row per sample-metric combination). This is needed for ggplot's facet_wrap().
# pivot_longer() stacks the three metric columns into two columns:
#   "metric" (which index) and "value" (the diversity score)
alpha_long <- alpha_div %>%
  tidyr::pivot_longer(
    cols      = c(Observed, Shannon, InvSimpson),  # Columns to stack
    names_to  = "metric",    # New column for the metric name
    values_to = "value"      # New column for the diversity value
  )
saveRDS(alpha_div, file.path(microbiome_rds_dir, "04_alpha_div.rds"))
saveRDS(alpha_long, file.path(microbiome_rds_dir, "05_alpha_long.rds"))
utils::write.csv(alpha_div, file.path(microbiome_tbl_dir, "01_alpha_diversity_table.csv"), row.names = FALSE)
utils::write.csv(alpha_long, file.path(microbiome_tbl_dir, "02_alpha_diversity_long.csv"), row.names = FALSE)
# Optional fast restart (uncomment if this step is too slow):
# alpha_div <- readRDS(file.path(microbiome_rds_dir, "04_alpha_div.rds"))
# alpha_long <- readRDS(file.path(microbiome_rds_dir, "05_alpha_long.rds"))

# Create a boxplot with individual data points overlaid for each diversity metric.
# facet_wrap() creates a separate panel for each metric, each with its own y-scale.
ggplot2::ggplot(alpha_long, ggplot2::aes(x = Treatment, y = value, fill = Treatment)) +
  ggplot2::geom_boxplot(width = 0.6, outlier.shape = NA, alpha = 0.7) +  # Box = median + quartiles
  ggplot2::geom_jitter(width = 0.15, size = 2, alpha = 0.6) +  # Individual samples as dots
  ggplot2::scale_fill_brewer(palette = "Dark2") +
  ggplot2::facet_wrap(~ metric, scales = "free_y") +  # Separate panel per metric
  ggplot2::theme_minimal(base_size = 13) +
  ggplot2::labs(
    x     = "Treatment",
    y     = "Diversity Value",
    title = "Alpha Diversity by Treatment",
    fill  = "Treatment"
  ) +
  ggplot2::theme(legend.position = "bottom")
```

### Statistical testing

We use **non-parametric tests** because diversity indices often have non-normal
distributions and our sample sizes are relatively small.

Test rationale in this section:

- **Kruskal-Wallis** is the omnibus test: "Does alpha diversity differ across
  treatment groups at all?"
- **Pairwise Wilcoxon** is the follow-up test: "Which specific group pairs
  differ?" (with FDR correction for multiple comparisons).

```{r alpha-stats}
# Kruskal-Wallis test: a non-parametric test for comparing 3+ groups.
# Unlike ANOVA, it does not assume the data follows a normal distribution —
# important because diversity indices are often skewed.
# The formula "metric ~ group" tests whether the metric differs across groups.
# A small p-value (< 0.05) indicates at least one group is significantly different.
alpha_div$Treatment <- droplevels(as.factor(alpha_div$Treatment))
for (metric in c("Observed", "Shannon", "InvSimpson")) {
  metric_df <- data.frame(
    value = alpha_div[[metric]],
    Treatment = alpha_div$Treatment
  )
  metric_df <- metric_df[stats::complete.cases(metric_df), , drop = FALSE]
  if (length(unique(metric_df$Treatment)) < 2) {
    cat(sprintf("%s: skipped (fewer than 2 treatment groups after filtering)\n", metric))
    next
  }
  kw <- stats::kruskal.test(
    value ~ Treatment, data = metric_df  # Test: diversity metric ~ treatment group
  )
  cat(sprintf("%s: Kruskal-Wallis chi-squared = %.2f, p = %.4f\n",
              metric, kw$statistic, kw$p.value))
}
```

```{r alpha-pairwise}
# If the Kruskal-Wallis test is significant, it tells us "at least one group
# differs" but not WHICH groups. Pairwise Wilcoxon tests compare every pair
# of groups (e.g., Group A vs. Group B, Group A vs. Group C, Group B vs. Group C).
# p.adjust.method = "fdr" corrects for multiple comparisons — when testing
# many pairs, some will appear significant by chance. FDR controls this.
cat("\nPairwise Wilcoxon tests for alpha diversity metrics (FDR-adjusted)\n")
if (length(unique(stats::na.omit(alpha_div$Treatment))) < 2) {
  cat("Skipping pairwise Wilcoxon: fewer than 2 treatment groups available.\n")
} else {
  alpha_metrics <- c("Observed", "Shannon", "InvSimpson")

  pw_tbl_list <- lapply(alpha_metrics, function(metric_name) {
    metric_df <- data.frame(
      value = alpha_div[[metric_name]],
      Treatment = alpha_div$Treatment
    )
    metric_df <- metric_df[stats::complete.cases(metric_df), , drop = FALSE]

    if (length(unique(metric_df$Treatment)) < 2) {
      return(NULL)
    }

    pw_metric <- stats::pairwise.wilcox.test(
      metric_df$value,
      metric_df$Treatment,
      p.adjust.method = "fdr"
    )

    pw_metric_tbl <- pw_metric$p.value %>%
      as.table() %>%
      as.data.frame(stringsAsFactors = FALSE) %>%
      dplyr::rename(
        group_1 = Var1,
        group_2 = Var2,
        p_value_fdr = Freq
      ) %>%
      dplyr::filter(!is.na(p_value_fdr)) %>%
      dplyr::mutate(metric = metric_name)

    if (nrow(pw_metric_tbl) == 0) {
      return(NULL)
    }

    pw_metric_tbl
  })

  pw_all_tbl <- dplyr::bind_rows(pw_tbl_list)

  if (nrow(pw_all_tbl) == 0) {
    cat("No valid pairwise comparisons after filtering.\n")
  } else {
    pw_all_tbl <- pw_all_tbl %>%
      dplyr::mutate(
        significant = dplyr::case_when(
          p_value_fdr < 0.001 ~ "***",
          p_value_fdr < 0.01  ~ "**",
          p_value_fdr < 0.05  ~ "*",
          TRUE                ~ "ns"
        ),
        p_value_fdr_display = dplyr::case_when(
          p_value_fdr == 0 ~ "< 1e-16",
          TRUE ~ formatC(p_value_fdr, format = "e", digits = 2)
        )
      ) %>%
      dplyr::arrange(metric, p_value_fdr) %>%
      dplyr::select(metric, group_1, group_2, p_value_fdr_display, significant)

    gt::gt(pw_all_tbl) %>%
      gt::cols_label(
        metric = "Metric",
        group_1 = "Group 1",
        group_2 = "Group 2",
        p_value_fdr_display = "FDR-adjusted p-value",
        significant = "Significance"
      ) %>%
      gt::tab_header(
        title = "Pairwise Wilcoxon Results",
        subtitle = "Observed, Shannon, and Inverse Simpson alpha diversity metrics"
      )
  }
}
```

> **Interpreting alpha diversity results:**
>
> - A significant difference in **Observed** richness means one group has
>   detectably more/fewer taxa.
> - A significant difference in **Shannon** indicates unequal evenness---one
>   group's community may be dominated by fewer taxa.
> - A significant difference in **Inverse Simpson** suggests differences in
>   dominance structure; lower values indicate communities dominated by fewer
>   taxa, while higher values indicate more even communities.
> - If a test is **not significant**, the data do not provide strong evidence
>   of a between-treatment difference for that metric (Observed, Shannon, or
>   Inverse Simpson) under the current sample size and variability.
> - Non-significant does **not** prove groups are identical---it can also
>   reflect small effect sizes, high within-group dispersion, or limited power.
> - Reduced alpha diversity is commonly observed in disease states ("dysbiosis")
>   but is not universal and should always be interpreted in the context of the
>   biological system.

> **For your own analyses, consider:**
>
> - These alpha-diversity tests are based on **relative abundance** data.
>   How might your conclusions change if you instead compared diversity using
>   **absolute abundance** (e.g., ASV counts scaled by `CopyNum_norm`)?
> - How might your results differ if diversity/statistics were computed at a
>   different taxonomic level (ASV vs Genus vs Family)?
> - Which biological question is best matched to each level of comparison?

> **Model choice note for your own analyses:**
>
> - The tests above assume samples are independent. If your design includes
>   repeated sampling of the same individual, technical/biological replicates,
>   or clustering within hives/nests, that assumption is violated.
> - In those cases, consider a **linear mixed-effects model** (or generalized
>   mixed model), with fixed effects for your treatment variables and random
>   effects for grouping factors (for example: bee ID, hive/nest, batch, or
>   sampling time structure).
> - Mixed-effects models are useful when you need to estimate treatment effects
>   while accounting for non-independence and hierarchical study design.

------------------------------------------------------------------------

## Beta Diversity: Between-Sample Diversity

**Beta diversity** measures how different microbial communities are *between*
samples. This involves two steps: (1) computing a distance matrix, and (2)
testing whether groups differ significantly.

### Distance metrics

| Metric | Data type | Considers |
|--------|-----------|-----------|
| **Bray-Curtis** | Abundance | Shared composition, weighted toward abundant taxa |
| **Canberra** | Abundance | Relative differences, with stronger influence from low-abundance taxa |
| **UniFrac (unweighted)** | Presence/absence + tree | Shared evolutionary history |
| **UniFrac (weighted)** | Abundance + tree | Abundance-weighted evolutionary history |

**Bray-Curtis** is widely used in amplicon studies. It generally puts more weight
on differences among abundant taxa.

**Canberra** emphasizes proportional differences in low-abundance/rarer taxa,
which can reveal shifts that are less visible with Bray-Curtis.

**UniFrac** incorporates phylogenetic relationships between taxa. Two communities
sharing closely related taxa will have lower UniFrac distance than two communities
sharing distantly related taxa, even if the number of shared taxa is the same.
Requires a phylogenetic tree in the phyloseq object.

### PCoA ordination

**Principal Coordinates Analysis (PCoA)** projects the high-dimensional distance
matrix onto 2--3 axes that capture the most variation. It is the standard
visualization for beta diversity.

- Points = samples
- Points closer together = more similar communities
- Axis labels show what percentage of total variation each axis captures

```{r pcoa-ordination, fig.height=5, fig.width=10}
# microViz workflow: keep ASV-level data ("unique"), then calculate distance + PCoA.
beta_ps <- bee_ps %>%
  microViz::tax_transform(trans = "identity", rank = "unique")

distance_map <- c(
  bray = "Bray-Curtis",
  canberra = "Canberra"
)

pcoa_plot_list <- lapply(names(distance_map), function(dist_name) {
  beta_ps %>%
    microViz::dist_calc(dist = dist_name) %>%
    microViz::ord_calc(method = "PCoA") %>%
    microViz::ord_plot(color = "Treatment", alpha = 0.8, size = 2.8) +
    ggplot2::stat_ellipse(ggplot2::aes(color = Treatment), type = "norm", linetype = 2, level = 0.95) +
    ggplot2::scale_color_brewer(palette = "Dark2") +
    ggplot2::theme_minimal(base_size = 12) +
    ggplot2::labs(title = paste0("PCoA: ", distance_map[[dist_name]], " Distance")) +
    ggplot2::theme(legend.position = "bottom")
})

saveRDS(
  list(beta_ps = beta_ps, distance_map = distance_map),
  file.path(microbiome_rds_dir, "06_beta_distance_setup.rds")
)

patchwork::wrap_plots(pcoa_plot_list, nrow = 1, guides = "collect")
```

> **Caution:** PCoA is a *visualization* tool, not a statistical test. Apparent
> visual separation must be confirmed with PERMANOVA (below). Human eyes are
> very good at finding patterns that may not be statistically significant.

### PERMANOVA: Testing for group differences

**PERMANOVA** (Permutational Multivariate Analysis of Variance) tests whether
group centroids (mean positions in multivariate space) differ significantly. It
works by:

Why this test here: beta diversity is a **distance matrix** (multivariate, not
single-response data), so PERMANOVA is appropriate for testing treatment-level
community composition differences in that space.

1. Computing a distance matrix (here: Bray-Curtis and Canberra)
2. Partitioning variation into between-group and within-group components
3. Permuting group labels thousands of times to generate a null distribution
4. Asking: is the observed between-group variation more extreme than expected
   by chance?

```{r permanova}
# Recreate setup if this chunk is run independently.
if (!exists("beta_ps") || !exists("distance_map")) {
  beta_ps <- bee_ps %>%
    microViz::tax_transform(trans = "identity", rank = "unique")
  distance_map <- c(
    bray = "Bray-Curtis",
    canberra = "Canberra"
  )
}

# Run PERMANOVA for each distance with microViz::dist_permanova.
permanova_tbl_list <- lapply(names(distance_map), function(dist_name) {
  perm_psx <- beta_ps %>%
    microViz::dist_calc(dist = dist_name) %>%
    microViz::dist_permanova(
      variables = "Treatment",
      n_perms = 9999,
      seed = 42,
      verbose = FALSE
    )

  perm_df <- as.data.frame(microViz::perm_get(perm_psx))
  term_row <- dplyr::case_when(
    "Treatment" %in% rownames(perm_df) ~ "Treatment",
    "Model" %in% rownames(perm_df) ~ "Model",
    TRUE ~ rownames(perm_df)[which(!is.na(perm_df$F))[1]]
  )

  data.frame(
    distance = distance_map[[dist_name]],
    term = term_row,
    r2 = perm_df[term_row, "R2"],
    pseudo_f = perm_df[term_row, "F"],
    p_value = perm_df[term_row, "Pr(>F)"],
    stringsAsFactors = FALSE
  )
})

permanova_tbl <- dplyr::bind_rows(permanova_tbl_list) %>%
  dplyr::mutate(
    p_value_display = dplyr::case_when(
      p_value == 0 ~ "< 1e-16",
      TRUE ~ formatC(p_value, format = "e", digits = 2)
    ),
    significant = dplyr::case_when(
      p_value < 0.001 ~ "***",
      p_value < 0.01 ~ "**",
      p_value < 0.05 ~ "*",
      TRUE ~ "ns"
    )
  )

saveRDS(permanova_tbl, file.path(microbiome_rds_dir, "07_permanova_results_bray_canberra.rds"))
utils::write.csv(permanova_tbl, file.path(microbiome_tbl_dir, "03_permanova_bray_canberra.csv"), row.names = FALSE)

permanova_tbl %>%
  dplyr::select(distance, r2, pseudo_f, p_value_display, significant) %>%
  gt::gt() %>%
  gt::fmt_number(columns = c(r2, pseudo_f), decimals = 3) %>%
  gt::cols_label(
    distance = "Distance Metric",
    r2 = "R²",
    pseudo_f = "Pseudo-F",
    p_value_display = "p-value",
    significant = "Significance"
  ) %>%
  gt::tab_header(
    title = "PERMANOVA Results by Distance Metric",
    subtitle = "microViz::dist_permanova with Treatment as predictor"
  )
```

Key values to report:

- **R^2^**: Proportion of total community variation explained by the grouping
  variable. R^2^ = 0.30 means bee type explains 30% of community variation.
- **Pseudo-F**: The test statistic (ratio of between-group to within-group
  variation). Larger = stronger group separation.
- **p-value**: Probability of observing this much group separation by chance.

### Betadisper: Checking the assumptions

PERMANOVA is sensitive to *both* location differences (centroids) and dispersion
differences (spread). A group that is simply more variable could produce a
significant PERMANOVA even if its centroid is in the same place.

Why this test here: **betadisper + permutest** checks whether within-group
dispersion is comparable across treatments, helping distinguish true centroid
shifts from differences in group spread.

`betadisper()` tests whether group dispersions (distances of samples to their
group centroid) are homogeneous. If dispersions are equal, the PERMANOVA result
can be confidently attributed to centroid differences.

```{r betadisper}
# Recreate setup if this chunk is run independently.
if (!exists("beta_ps") || !exists("distance_map")) {
  beta_ps <- bee_ps %>%
    microViz::tax_transform(trans = "identity", rank = "unique")
  distance_map <- c(
    bray = "Bray-Curtis",
    canberra = "Canberra"
  )
}

# microViz wrapper for betadisper on each distance matrix.
dispersion_tbl_list <- lapply(names(distance_map), function(dist_name) {
  bdisp_psx <- beta_ps %>%
    microViz::dist_calc(dist = dist_name) %>%
    microViz::dist_bdisp(variables = "Treatment", verbose = FALSE)

  bdisp_obj <- microViz::bdisp_get(bdisp_psx)$Treatment
  anova_df <- as.data.frame(bdisp_obj$anova)
  group_row <- rownames(anova_df)[1]
  group_means <- tapply(
    X = bdisp_obj$model$distances,
    INDEX = bdisp_obj$model$group,
    FUN = mean,
    na.rm = TRUE
  )
  group_means_text <- paste(
    paste0(names(group_means), ": ", formatC(group_means, format = "f", digits = 3)),
    collapse = "; "
  )

  data.frame(
    distance = distance_map[[dist_name]],
    f_value = anova_df[group_row, "F value"],
    p_value = anova_df[group_row, "Pr(>F)"],
    group_mean_distance = group_means_text,
    stringsAsFactors = FALSE
  )
})

dispersion_tbl <- dplyr::bind_rows(dispersion_tbl_list) %>%
  dplyr::mutate(
    p_value_display = dplyr::case_when(
      p_value == 0 ~ "< 1e-16",
      TRUE ~ formatC(p_value, format = "e", digits = 2)
    ),
    significant = dplyr::case_when(
      p_value < 0.001 ~ "***",
      p_value < 0.01 ~ "**",
      p_value < 0.05 ~ "*",
      TRUE ~ "ns"
    )
  )

saveRDS(dispersion_tbl, file.path(microbiome_rds_dir, "08_dispersion_results_bray_canberra.rds"))
utils::write.csv(dispersion_tbl, file.path(microbiome_tbl_dir, "04_betadisper_bray_canberra.csv"), row.names = FALSE)

dispersion_tbl %>%
  dplyr::select(distance, f_value, p_value_display, significant, group_mean_distance) %>%
  gt::gt() %>%
  gt::fmt_number(columns = f_value, decimals = 3) %>%
  gt::cols_label(
    distance = "Distance Metric",
    f_value = "Dispersion F",
    p_value_display = "p-value",
    significant = "Significance",
    group_mean_distance = "Mean Distance to Group Centroid"
  ) %>%
  gt::tab_header(
    title = "Dispersion (betadisper) Results by Distance Metric",
    subtitle = "microViz::dist_bdisp on Treatment groups"
  )
```

> **Interpretation framework:**
>
> - PERMANOVA significant + betadisper *not* significant → Groups have
>   **different community compositions** (true centroid difference).
> - PERMANOVA significant + betadisper significant → Interpret with caution.
>   Groups may differ in variability rather than (or in addition to) composition.
> - PERMANOVA *not* significant → No evidence of compositional differences
>   between groups.

```{r beta-interpretation}
# Summarize PERMANOVA and dispersion checks for each distance metric.
cat("\n=== Beta Diversity Summary ===\n")
for (i in seq_len(nrow(permanova_tbl))) {
  perm_row <- permanova_tbl[i, , drop = FALSE]
  disp_row <- dispersion_tbl[dispersion_tbl$distance == perm_row$distance, , drop = FALSE]
  cat(sprintf(
    "%s: Treatment explains %.1f%% of variation (R² = %.3f, pseudo-F = %.2f, p = %.4g). ",
    perm_row$distance,
    perm_row$r2 * 100,
    perm_row$r2,
    perm_row$pseudo_f,
    perm_row$p_value
  ))
  if (nrow(disp_row) == 1) {
    cat(sprintf(
      "Dispersion test: F = %.2f, p = %.4g.\n",
      disp_row$f_value,
      disp_row$p_value
    ))
  } else {
    cat("Dispersion test not available.\n")
  }
}
```

If results differ between Bray-Curtis and Canberra, consider *which part of the
community is changing*: Bray-Curtis differences usually point to shifts in
dominant taxa, while Canberra differences often highlight changes in rarer taxa.
Ecologically, disagreement between metrics can indicate that treatment effects
are concentrated in either the core community or the rare biosphere, rather
than uniformly across all taxa.


------------------------------------------------------------------------

## Differential Abundance: Which Taxa Differ?

Alpha and beta diversity tell us that communities differ *overall*. Differential
abundance (DA) analysis identifies the **specific taxa** driving those
differences---the biologically actionable output.

### Setup and prevalence filtering

Before testing hundreds of taxa, we remove low-prevalence ASVs (present in
<10% of samples). This reduces noise and the multiple-testing burden.

```{r da-setup}
# For differential abundance testing, we need the data in a flat table format
# (one big table with both metadata and ASV counts) rather than a phyloseq object.

# Extract the ASV count matrix from phyloseq and convert to a plain R matrix.
# as(..., "matrix") coerces the phyloseq otu_table into a standard matrix.
otu_mat <- as(phyloseq::otu_table(bee_ps), "matrix")

# Ensure samples are rows and ASVs are columns (transpose if needed)
if (phyloseq::taxa_are_rows(bee_ps)) otu_mat <- t(otu_mat)

# Convert to data.table format for fast operations.
# keep.rownames = "Sample" preserves the sample names as a column called "Sample".
otu_dt   <- data.table::as.data.table(otu_mat, keep.rownames = "Sample")
meta_dt  <- data.table::as.data.table(
  as(phyloseq::sample_data(bee_ps), "data.frame"),  # Extract metadata as data frame
  keep.rownames = "Sample"                            # Keep sample names as a column
)
taxa_ids <- phyloseq::taxa_names(bee_ps)  # All ASV names in the dataset

# Join metadata and ASV counts into a single wide table (one row per sample).
# on = "Sample" tells data.table to merge on the shared "Sample" column.
full_dt <- meta_dt[otu_dt, on = "Sample"]

# Prevalence filter: remove rare ASVs before testing to reduce noise.
# colSums(otu_mat > 0) counts how many samples each ASV appears in (presence/absence).
# Dividing by n_samples gives the fraction of samples (0 to 1).
# We keep only ASVs present in at least 10% of samples.
n_samples <- nrow(otu_dt)
prevalence_pct <- colSums(otu_mat > 0) / n_samples
keep_for_da <- names(prevalence_pct[prevalence_pct >= 0.10])
saveRDS(
  list(otu_mat = otu_mat, otu_dt = otu_dt, meta_dt = meta_dt, full_dt = full_dt, keep_for_da = keep_for_da),
  file.path(microbiome_rds_dir, "10_da_setup_objects.rds")
)
# Optional fast restart (uncomment if this step is too slow):
# da_setup <- readRDS(file.path(microbiome_rds_dir, "10_da_setup_objects.rds"))
# otu_mat <- da_setup$otu_mat; otu_dt <- da_setup$otu_dt; meta_dt <- da_setup$meta_dt
# full_dt <- da_setup$full_dt; keep_for_da <- da_setup$keep_for_da

cat("ASVs before prevalence filter:", length(taxa_ids), "\n")
cat("ASVs after prevalence filter (>= 10%):", length(keep_for_da), "\n")
```

### DESeq2 differential abundance per taxon

For each taxon, we test whether abundance differs between groups using
**DESeq2** (negative binomial GLM). This approach is generally preferred for
amplicon count data because it models mean-variance relationships explicitly and
handles library-size normalization internally.

Why this test here: this section performs many **two-group, per-taxon**
comparisons on count data; DESeq2 is robust for overdispersed counts and
returns effect sizes (log2 fold changes) with FDR-adjusted p-values.

```{r da-deseq2}
# For a clean two-group comparison, subset to two groups of interest.
# Here we compare the first two treatment levels alphabetically.
treatment_levels <- sort(unique(as.character(full_dt$Treatment)))
if (length(treatment_levels) < 2) {
  stop("Need at least two Treatment levels for differential abundance testing.")
}
comparison_groups <- treatment_levels[1:2]
cat("DA comparison groups:", paste(comparison_groups, collapse = " vs "), "\n")
da_dt <- full_dt[Treatment %in% comparison_groups]

if (!requireNamespace("DESeq2", quietly = TRUE)) {
  stop("Package 'DESeq2' is required for this section. Install via BiocManager::install('DESeq2').")
}

# Build a phyloseq object for the two-group comparison and prevalence-filtered taxa.
ps_da <- phyloseq::prune_samples(
  as.character(phyloseq::sample_data(bee_ps)$Treatment) %in% comparison_groups,
  bee_ps
)
ps_da <- phyloseq::prune_taxa(phyloseq::taxa_names(ps_da) %in% keep_for_da, ps_da)

# Ensure Treatment is an ordered factor so the contrast direction is explicit.
ps_da_meta <- data.frame(phyloseq::sample_data(ps_da), check.names = FALSE)
ps_da_meta$Treatment <- factor(as.character(ps_da_meta$Treatment), levels = comparison_groups)
phyloseq::sample_data(ps_da) <- phyloseq::sample_data(ps_da_meta)

# Convert to DESeq2 and run Wald tests with settings commonly used for
# zero-heavy amplicon count data.
deseq_dds <- phyloseq::phyloseq_to_deseq2(ps_da, ~ Treatment)
deseq_dds <- DESeq2::DESeq(
  deseq_dds,
  test = "Wald",
  fitType = "local",
  sfType = "poscounts",
  quiet = TRUE
)

# Contrast direction: positive log2FC means enriched in comparison_groups[2]
# relative to comparison_groups[1].
deseq_res <- DESeq2::results(
  deseq_dds,
  contrast = c("Treatment", comparison_groups[2], comparison_groups[1]),
  alpha = 0.05
)

deseq_df <- as.data.frame(deseq_res) %>%
  tibble::rownames_to_column("taxon") %>%
  dplyr::rename(p_value = pvalue, p_adj = padj) %>%
  dplyr::arrange(p_adj)

# Significant taxa after FDR correction.
sig_taxa <- deseq_df %>%
  dplyr::filter(!is.na(p_adj), p_adj <= 0.05)

cat("Significant taxa (DESeq2 FDR <= 0.05):", nrow(sig_taxa), "of", nrow(deseq_df), "\n\n")
saveRDS(
  list(
    deseq_dds = deseq_dds,
    deseq_df = deseq_df,
    sig_taxa = sig_taxa,
    comparison_groups = comparison_groups,
    da_dt = da_dt
  ),
  file.path(microbiome_rds_dir, "11_da_deseq2_results.rds")
)
utils::write.csv(deseq_df, file.path(microbiome_tbl_dir, "04_da_deseq2_all_taxa.csv"), row.names = FALSE)
utils::write.csv(sig_taxa, file.path(microbiome_tbl_dir, "05_da_significant_taxa.csv"), row.names = FALSE)
# Optional fast restart (uncomment if this step is too slow):
# da_deseq2 <- readRDS(file.path(microbiome_rds_dir, "11_da_deseq2_results.rds"))
# deseq_dds <- da_deseq2$deseq_dds; deseq_df <- da_deseq2$deseq_df
# sig_taxa <- da_deseq2$sig_taxa; comparison_groups <- da_deseq2$comparison_groups
# da_dt <- da_deseq2$da_dt

# For the significant ASVs, look up their taxonomy (Family, Genus) so we
# know WHAT organisms are driving the difference, not just ASV IDs.
if (nrow(sig_taxa) > 0) {
  tax_info <- as.data.frame(phyloseq::tax_table(bee_ps))  # Full taxonomy table
  sig_taxa_info <- sig_taxa %>%
    dplyr::left_join(
      tax_info %>% tibble::rownames_to_column("taxon"),  # Add "taxon" column for joining
      by = "taxon"
    ) %>%
    dplyr::select(taxon, Family, Genus, log2FoldChange, lfcSE, p_value, p_adj) %>%  # Keep relevant columns
    dplyr::arrange(p_adj)  # Sort by significance (most significant first)
  utils::write.csv(
    sig_taxa_info,
    file.path(microbiome_tbl_dir, "06_da_significant_taxa_with_taxonomy.csv"),
    row.names = FALSE
  )
  # Reuse the same genus palette used in composition barplots.
  genus_color_map <- genus_palette
  default_genus_color <- unname(genus_color_map["Other"])
  if (length(default_genus_color) == 0 || is.na(default_genus_color)) {
    default_genus_color <- "grey90"
  }

  sig_taxa_gt_df <- sig_taxa_info %>%
    dplyr::mutate(
      p_value_display = dplyr::case_when(
        p_value == 0 ~ "< 1e-16",
        TRUE ~ formatC(p_value, format = "e", digits = 2)
      ),
      p_adj_display = dplyr::case_when(
        p_adj == 0 ~ "< 1e-16",
        TRUE ~ formatC(p_adj, format = "e", digits = 2)
      ),
      Genus_color = dplyr::case_when(
        !is.na(Genus) & Genus %in% names(genus_color_map) ~ unname(genus_color_map[Genus]),
        TRUE ~ default_genus_color
      )
    ) %>%
    dplyr::select(
      taxon, Family, Genus, log2FoldChange, lfcSE, p_value_display, p_adj_display, Genus_color
    )

  gt_tbl <- sig_taxa_gt_df %>%
    gt::gt() %>%
    gt::cols_label(
      taxon = "Taxon",
      Family = "Family",
      Genus = "Genus",
      log2FoldChange = "log2 Fold Change",
      lfcSE = "LFC SE",
      p_value_display = "Raw p-value",
      p_adj_display = "FDR-adjusted p-value"
    ) %>%
    gt::data_color(
      columns = Genus,
      fn = function(x) {
        dplyr::case_when(
          !is.na(x) & x %in% names(genus_color_map) ~ unname(genus_color_map[x]),
          TRUE ~ default_genus_color
        )
      },
      apply_to = "fill",
      autocolor_text = TRUE
    ) %>%
    gt::cols_hide(columns = Genus_color) %>%
    gt::tab_header(
      title = "Significant Differentially Abundant Taxa",
      subtitle = paste0(
        "DESeq2 Wald tests (",
        comparison_groups[2],
        " vs ",
        comparison_groups[1],
        "); positive log2FC = enriched in ",
        comparison_groups[2]
      )
    )

  gt_tbl
}
```

### Visualize differentially abundant taxa

This figure shows sample-level abundances for **DESeq2-significant ASVs**.  
Each panel is one significant ASV, points are individual samples, and the red
error bars indicate the bootstrapped mean and 95% CI within each treatment.
Values are DESeq2 size-factor-normalized counts, so panels are consistent with
the differential abundance model.

```{r da-visualization, fig.height=6, fig.width=10}
if (nrow(sig_taxa) > 0) {
  # Get the list of significant ASV names to plot
  plot_taxa <- sig_taxa$taxon

  # Use DESeq2 size-factor normalized counts so visualization is consistent
  # with the DESeq2 differential abundance model.
  norm_counts <- DESeq2::counts(deseq_dds, normalized = TRUE)
  norm_counts <- norm_counts[plot_taxa, , drop = FALSE]

  # Convert normalized count matrix (taxa x samples) to long format.
  # One row = one sample-taxon abundance observation.
  norm_counts_long <- as.data.frame(norm_counts) %>%
    tibble::rownames_to_column("taxon") %>%
    tidyr::pivot_longer(
      cols      = -taxon,
      names_to  = "Sample",
      values_to = "norm_abundance"
    )

  # Sample-level metadata needed for grouping on the x-axis.
  sample_groups <- data.frame(phyloseq::sample_data(ps_da), check.names = FALSE) %>%
    tibble::rownames_to_column("Sample") %>%
    dplyr::select(Sample, Treatment)

  # Create unique plotting labels per ASV so taxa are not collapsed when several
  # ASVs share the same genus/family label.
  tax_labels <- as.data.frame(phyloseq::tax_table(bee_ps)) %>%
    tibble::rownames_to_column("taxon") %>%
    dplyr::mutate(
      tax_label = dplyr::coalesce(as.character(Genus), as.character(Family), taxon),
      label = paste0(taxon, " | ", tax_label)
    ) %>%
    dplyr::select(taxon, tax_label, label)

  # Merge sample metadata and taxonomy labels into plotting data.
  da_plot_dt <- norm_counts_long %>%
    dplyr::left_join(sample_groups, by = "Sample") %>%
    dplyr::left_join(tax_labels, by = "taxon")
  saveRDS(da_plot_dt, file.path(microbiome_rds_dir, "12_da_plot_dt.rds"))
  # Optional fast restart (uncomment if this step is too slow):
  # da_plot_dt <- readRDS(file.path(microbiome_rds_dir, "12_da_plot_dt.rds"))

  # Create a jitter plot with one panel per significant ASV.
  # y uses DESeq2 normalized counts for consistency with the DA model.
  ggplot2::ggplot(da_plot_dt, ggplot2::aes(x = Treatment, y = norm_abundance, color = Treatment)) +
    ggplot2::geom_jitter(width = 0.2, alpha = 0.6, size = 1.5) +  # Individual sample dots
    ggplot2::stat_summary(
      fun.data = "mean_se",       # Mean +/- standard error (no extra package dependency)
      color    = "red",
      geom     = "errorbar",      # Display as error bars
      width    = 0.5
    ) +
    ggplot2::scale_color_brewer(palette = "Dark2") +
    ggplot2::facet_wrap(~ label, scales = "free_y") +  # One panel per ASV, independent y-axes
    ggplot2::theme_minimal(base_size = 11) +
    ggplot2::labs(
      x     = "Treatment",
      y     = "DESeq2 normalized abundance",
      title = paste0(
        "DESeq2 Significant Taxa: ",
        comparison_groups[2],
        " vs ",
        comparison_groups[1]
      )
    ) +
    ggplot2::theme(strip.text = ggplot2::element_text(size = 8))  # Smaller panel titles
}
```

### Heatmap of differential taxa

This heatmap summarizes **only the significant ASVs** from DESeq2.  
Each row is one significant ASV and each column is a treatment group (CL, MD).  
Tile color shows the **mean DESeq2-normalized abundance** of that ASV within each
treatment, displayed on a `log(mean + 1)` scale to make both low- and
high-abundance taxa easier to compare visually.

```{r da-heatmap, fig.height=5, fig.width=6}
if (nrow(sig_taxa) > 0) {
  # Calculate mean DESeq2-normalized abundance of each significant ASV
  # within each treatment. Keep ASV-resolved labels to match the DESeq2 table.
  # group_by() splits the data by taxon label and treatment; summarize() computes
  # the mean for each group. .groups = "drop" removes the grouping afterward.
  heatmap_dt <- da_plot_dt %>%
    dplyr::group_by(taxon, label, tax_label, Treatment) %>%
    dplyr::summarize(mean_abund = mean(norm_abundance), .groups = "drop")

  # Order rows by overall abundance (across both treatments), not alphabetically.
  # Highest-abundance taxa are shown at the top of the y-axis.
  label_levels <- heatmap_dt %>%
    dplyr::group_by(label) %>%
    dplyr::summarise(overall_mean_abund = mean(mean_abund, na.rm = TRUE), .groups = "drop") %>%
    dplyr::arrange(dplyr::desc(overall_mean_abund)) %>%
    dplyr::pull(label)

  heatmap_dt <- heatmap_dt %>%
    dplyr::mutate(label = factor(label, levels = rev(label_levels)))
  saveRDS(heatmap_dt, file.path(microbiome_rds_dir, "13_heatmap_dt.rds"))
  # Optional fast restart (uncomment if this step is too slow):
  # heatmap_dt <- readRDS(file.path(microbiome_rds_dir, "13_heatmap_dt.rds"))

  # Create a heatmap: rows = taxa, columns = treatments, color = mean
  # DESeq2-normalized abundance.
  # geom_tile() draws colored rectangles; log1p() applies log(x + 1) to compress
  # the range (microbiome counts span orders of magnitude, so log-scaling
  # prevents a few high-abundance taxa from washing out all the color variation).
  # Because log1p(mean_abund) is always >= 0, use a sequential (not diverging)
  # color scale.
  ggplot2::ggplot(heatmap_dt, ggplot2::aes(x = Treatment, y = label)) +
    ggplot2::geom_tile(ggplot2::aes(fill = log1p(mean_abund)), color = "black") +  # Colored tiles with black borders
    ggplot2::scale_fill_gradient(
      low  = "grey95",         # Low abundance
      high = "steelblue4",     # High abundance
      name = "log(mean + 1)"   # Legend title
    ) +
    ggplot2::theme_minimal(base_size = 12) +
    ggplot2::labs(
      x     = "Treatment",
      y     = "",
      title = "Mean DESeq2-Normalized Abundance of Significant ASVs"
    ) +
    ggplot2::coord_fixed(ratio = 0.5)  # Keep tiles roughly square-shaped
}
```

> **A note on DA methods:** There are many approaches to differential abundance
> analysis, and different methods can yield different results (Nearing et al.,
> 2022, *Nature Communications*). Other commonly used tools include:
>
> - **DESeq2**: Negative binomial GLM; handles normalization internally; good
>   for complex experimental designs.
> - **ANCOM-BC**: Explicitly addresses compositionality of microbiome data.
> - **MaAsLin2**: Flexible for multivariable models and mixed effects.
>
> Best practice: run 2+ methods and focus on concordant results. Always evaluate
> effect size and biological plausibility alongside statistical significance.

------------------------------------------------------------------------

## Wrap-Up: The Complete Picture

We have walked through the full microbiome analysis pipeline:

```
Raw reads → Quality filtering → DADA2 denoising → ASV table → Taxonomy
    → Phyloseq → Composition → Alpha diversity → Beta diversity → DA
```

**Each analysis answers a different biological question:**

| Analysis | Question | Key Output |
|----------|----------|------------|
| Composition plots | What taxa are present and at what proportions? | Stacked barplots |
| Alpha diversity | How diverse is each sample, and do groups differ? | Shannon, Observed, tests |
| Beta diversity + PERMANOVA | Do overall community structures differ? | PCoA, R², p-value |
| Differential abundance | Which specific taxa drive the differences? | Significant taxa list |

### Resources for further learning

- **DADA2 tutorial**: <https://benjjneb.github.io/dada2/tutorial.html>
- **phyloseq documentation**: <https://joey711.github.io/phyloseq/>
- **microViz package**: <https://david-barnett.github.io/microViz/>
- **DA method comparison** (Nearing et al., 2022): <https://doi.org/10.1038/s41467-022-28034-z>
- **SILVA database**: <https://www.arb-silva.de/>

------------------------------------------------------------------------

*Workshop tutorial created by Michael Sieler. Last updated: `r Sys.Date()`.*
